---
format:
  html:
    self-contained: true 
    toc: true
    toc-depth: 3     
    number-sections: true
  pdf:
    fig-format: png
    toc: true
    toc-depth: 3  
    number-sections: true
    execute:
      echo: false # don't show code in PDF
execute:
  echo: true # show code (by default for HTML)
  warning: false # don't show warnings
  error: false # only render if no errors arise
  message: false
  freeze: auto # prevent recomputation unless code changes
bibliography: references.bib 
csl: https://www.zotero.org/styles/apa # citation style
knitr:
  opts_knit:
    root.dir: "~/Documents/00_ESDSRS/semester1/3_soil/soil-N-project" # working directory
---



# Abstract

This study evaluates the spatial prediction of soil C:N ratios across Germany by comparing random forest and cubist models. Soil data from the LUCAS 2015 survey (n = 1690) were combined with a harmonized set of environmental predictors representing terrain, climate, soil properties, land cover and remote sensing variables. Model performance was assessed using 10-fold cross-validation, while spatial predictions, uncertainty and model disagreement were analysed using a 30-iteration bootstrapping approach.
Both models achieved comparable predictive performance (RMSE ≈ 2.7; R² ≈ 0.60; CCC ≈ 0.75), indicating similar overall accuracy. However, they produced consistent spatial differences, particularly in heterogeneous and topographically complex regions. Random forest showed lower overall uncertainty, while both models compressed the predicted C:N range. Predicted values were slightly higher in Natura 2000 grasslands, although effect sizes remained small.
Overall, the findings demonstrate that similar model performance does not imply identical spatial predictions and highlight the need to account for model-dependent variation in digital soil mapping.

{{< pagebreak >}}

# Introduction

The soil carbon-to-nitrogen (C:N) ratio is a widely used indicator of soil biogeochemical functioning because it reflects the balance between carbon and nitrogen in soil organic matter and is closely linked to nutrient cycling processes [@cleveland_cnp_2007; @sardans_cnp_2012]. Changes in soil C:N ratios influence microbial nutrient demand, decomposition dynamics and nitrogen mineralization, thereby shaping soil organic matter turnover and broader ecosystem functioning [@cleveland_cnp_2007; @springob_bulk_2003]. For this reason, the C:N ratio is frequently used as an indicator of soil organic matter quality and nutrient availability across terrestrial ecosystems [@ostrowska_assessment_2015; @springob_bulk_2003]. Understanding how soil C:N ratios vary across space is therefore essential for analysing large-scale patterns of soil functioning and nutrient cycling.

Spatial modeling provides a framework to analyse such variation beyond sampled locations by linking measured soil properties with spatially explicit environmental covariates representing soil-forming factors and environmental gradients [@mcbratney_digital_2003; @wadoux_machine_2020]. In the context of soil stoichiometry, spatial prediction enables the identification of relationships between C:N ratios and environmental drivers such as climate, topography, soil properties and land use [@hong_path_2026; @sardans_cnp_2012]. Mapping these relationships allows the investigation of how environmental gradients shape spatial patterns of soil carbon and nitrogen dynamics across landscapes and regions.

Spatial prediction of soil C:N ratios typically relies on laboratory-derived soil observations from georeferenced sampling sites that serve as ground-truth data for model calibration and validation within digital soil mapping approaches [@mcbratney_digital_2003; @wadoux_machine_2020]. The LUCAS soil survey represents one of the most widely used datasets for large-scale soil modeling in Europe, providing harmonised measurements of soil organic carbon, total nitrogen and related soil properties across the European Union [@orgiazzi_lucas_2018; @hong_path_2026]. These observations are commonly combined with environmental covariates representing major environmental gradients controlling soil carbon and nitrogen dynamics, including terrain attributes derived from digital elevation models, climatic variables, soil maps, land-use layers and remotely sensed predictors such as vegetation indices or radar observations [@wadoux_machine_2020; @khaledian_selecting_2020]. The increasing integration of remote sensing data into soil modeling is further illustrated by recent continental-scale studies combining LUCAS observations with multi-source satellite data to predict spatial patterns of soil C:N ratios [@wang_mapping_2025].

Using such datasets, an increasing number of studies have applied spatial modeling approaches to predict soil C:N ratios across regional to continental scales [@hong_path_2026; @zhou_prediction_2021; @khosravani_soil_2026]. These studies consistently show that spatial variability in soil C:N ratios emerges from interactions among multiple environmental gradients rather than from a single dominant predictor. However, the relative importance of individual predictors varies considerably depending on environmental conditions, predictor selection and modeling strategy [@hong_path_2026; @zhou_prediction_2021; @khaledian_selecting_2020]. In addition, predictor importance may vary with spatial scale, with broad-scale studies emphasizing climatic gradients and large-scale structures, whereas finer-scale analyses highlight local soil properties and land-management effects [@minasny_digital_2016; @wadoux_machine_2020].

At the same time, the geographical distribution of spatial modeling studies on soil stoichiometry remains uneven. A substantial proportion of studies has been conducted in East Asia, particularly in China, whereas comparatively few studies exist for Europe [@hong_path_2026]. In Germany, national-scale modeling of soil C:N ratios remains limited to a small number of recent studies [@khosravani_soil_2026; @sakhaee_spatial_2024]. In addition, studies on spatial modeling of soil C:N ratios remain unevenly distributed across land-use systems [@deng_spatial_2020; @hong_path_2026; @sakhaee_spatial_2024; @zhou_prediction_2021]. Many analyses focus on cropland or mixed land-use datasets, whereas grassland soils are comparatively underrepresented. This is particularly noteworthy given the importance of grasslands for terrestrial carbon storage and their sensitivity to land-use-related variation in soil properties [@herold_soil_2014]. In grassland systems, land-use intensity can be characterized through well-defined management practices such as fertilisation, grazing and mowing [@bluthgen_quantitative_2012], which influence nutrient inputs, biomass removal and redistribution processes. These management differences are associated with measurable ecosystem responses under contrasting management regimes [@schreiner_slow_2026]. Consequently, grasslands provide a suitable study system for evaluating whether predicted spatial patterns of soil C:N ratios reflect known effects of land use and management. 

To capture such relationships, a range of statistical and machine-learning algorithms has been applied in digital soil mapping. Earlier approaches often relied on regression-based models or geostatistical interpolation techniques such as regression kriging, whereas more recent studies increasingly use machine-learning algorithms capable of modeling nonlinear relationships between environmental predictors and soil properties [@khaledian_selecting_2020; @wadoux_machine_2020]. Commonly applied models include random forest, cubist and other ensemble approaches [@khaledian_selecting_2020].

However, the methodological basis for selecting specific modeling algorithms often remains insufficiently transparent, even though different models may produce different prediction accuracies, variable importance structures and spatial prediction patterns when applied to identical datasets. Reported predictive performance in digital soil mapping varies widely depending on the target variable, spatial scale, predictor set and modeling approach, and is typically described as ranging from moderate to high explanatory power [@khaledian_selecting_2020; @wadoux_machine_2020]. 

In addition, modeling C:N ratios poses a specific methodological challenge because the target variable represents a ratio derived from two independently measured soil properties. Uncertainty in both carbon and nitrogen measurements may propagate into the resulting ratio and thereby influence model outcomes and their interpretation. Recent reviews indicate that soil C:N ratios are predominantly modeled directly, although alternative modeling strategies also occur [@hong_path_2026].

Beyond statistical performance, interpreting spatial prediction maps requires evaluating ecological plausibility and prediction uncertainty. Previous work has highlighted that digital soil mapping products may contain systematic biases or spatial uncertainties that are not always explicitly assessed [@hong_path_2026]. Therefore, interpretation increasingly involves comparison with known environmental gradients or ecological processes [@wadoux_machine_2020].

Against this background, important gaps remain. Systematic comparisons of machine-learning algorithms for predicting soil C:N ratios at national scale are still limited, particularly in Europe. In addition, the ecological plausibility of predicted patterns is not consistently evaluated, and grassland systems remain underrepresented as independent validation contexts.

The present study therefore investigates the spatial prediction of soil C:N ratios across Germany using two machine learning algorithms. Soil C:N ratios derived from the LUCAS soil database were modelled using random forest and cubist machine learning based on a harmonized set of 15 environmental predictors. The resulting prediction maps were compared with respect to model performance, variable importance, as well as spatial prediction patterns. In addition, the ecological plausibility of predicted patterns was assessed by comparing modeled C:N ratios of grassland soils within and outside Natura 2000 protected areas as a proxy for management intensity.


{{< pagebreak >}}

# Methods

The analysis was conducted using R statistical software [v4.5.2; @r_cite] integrated in Quarto [@Allaire_Quarto_2025]. We extensively used the packages terra [@terra], caret [@caret], and tidyverse [@tidyverse].

```{r}
#| label: setup
#| include: false

library(glcm)
library(Cubist)
library(raster) # required for dissimilarity
library(randomForest)
library(caret) # for k-fold cross-validation
library(sf)
library(gstat)
library(effsize)
library(terra)
library(tmap)
library(devtools)
library(geodata)
library(spatstat) # for computing distance between sampling points
library(ggplot2)
library(patchwork)
library(tidyterra)
library(tidyverse)
library(glue)

# don't show terra progress bar
terra::terraOptions(progress = 0)

# function for displaying raster summary as mean ± sd
show_mean_sd <- function(raster) {
  raster_val <- values(raster)
  mean_val <- mean(raster_val, na.rm = TRUE)
  sd_val <- sd(raster_val, na.rm = TRUE)
  return(paste(round(mean_val, 2), '±', round(sd_val,2)))
}
```

The prediction of soil C:N was conducted on the scale of Germany. The shapefile of the administrative boundaries was obtained from `rnaturalearth` [@rnaturalearth].

```{r}
#| label: load-germany-mask

library(rnaturalearth)
library(devtools)
# devtools::install_github("ropensci/rnaturalearthhires", quiet = TRUE)

# get Germany boundary polygon
germany <- ne_countries(scale = 10, country = "Germany", returnclass = "sf") %>%
  select(geometry)

# reproject to UTM 32N
crs <- 'EPSG:25832'
germany_4326 <- germany
germany <- st_transform(germany, crs)

```


## C:N ratio

```{r}
#| label: load-lucas-data 

# load LUCAS topsoil data
lucas_eu <- suppressMessages(
  sf::st_read('data/LUCAS2015_topsoildata_20200323/LUCAS_Topsoil_2015_20200323-shapefile/LUCAS_Topsoil_2015_20200323.shp', quiet = TRUE)
)

# reproject
lucas_eu <- st_transform(lucas_eu, crs)

# subset to Germany
lucas <- st_filter(lucas_eu, germany)
```

```{r}
#| label: calculate-sampling-point-distance

# prepare point pattern with projection (required for distance)
lucas_ppp <- as.ppp(lucas)

# find distance to nearest neighbor for each point
distances <- nndist(lucas_ppp)

# summary statistics (in km2)
mean_dist <- mean(distances)/1000
sd_dist <- sd(distances)/1000
min_dist <- min(distances)/1000
max_dist <- max(distances)/1000
```

The measurements of soil properties were obtained from the LUCAS topsoil survey 2015 [@jones_arwyn_lucas_2020]. In Germany, it consisted of `r nrow(lucas)` samples. Sampling points had a mean distance of `r round(mean_dist,2)` ± `r round(sd_dist,2)`, with minimum and maximum distances of `r round(min_dist,2)` and `r round(max_dist,2)`, respectively. This data was used to calculate the C:N ratios by dividing soil organic carbon (OC) by total nitrogen (N).

```{r}
#| label: fig-cn-distribution
#| fig-cap: "Distribution of C:N ratio values derived from LUCAS 2015 samples in Germany."
#| fig-height: 3
#| fig-dpi: 200

# first make columns numeric, then add new column with ratio
lucas <- lucas %>%
  mutate(N = if_else(N == "< LOD", 0, as.numeric(N)), 
         OC = if_else(OC == "< LOD", 0, as.numeric(OC)),
         cn_ratio = OC / N)

# summary statistics
cn_mean <- mean(lucas$cn_ratio)
cn_median <- median(lucas$cn_ratio)
cn_sd <- sd(lucas$cn_ratio)

ggplot(lucas, aes(cn_ratio)) +
  geom_histogram(binwidth = 0.5) +
  labs(x = "C:N ratio", y = "number of samples") +
  annotate(geom = "text", 
           x = 22, 
           y = 190, 
           label = paste('mean:', round(cn_mean,2)),
           hjust = 0, vjust = 1) + 
  annotate(geom = "text", 
           x = 22, 
           y = 165, 
           label = paste('median:', round(cn_median,2)),
           hjust = 0, vjust = 1) +
  annotate(geom = "text", 
           x = 22, 
           y = 140, 
           label = paste('standard deviation:', round(cn_sd,2)),
           hjust = 0, vjust = 1)

```

The distribution of C:N ratios obtained this way was strongly right-skewed, with most values below 10 (@fig-cn-distribution). There was one notable outlier at a ratio of almost 40, but it likely did not arise from measuring error since a comparable value can also be seen in the later LUCAS 2018 survey.

## Covariates

The selection of environmental predictors was largely based on those proposed by @khosravani_soil_2026, who had identified them according to expert knowledge. They were slightly modified where accurate replication was not possible due to lacking documentation in the paper or data availability. 

Elevation data at a resolution of 250 m was obtained from the global Digital Elevation Model provided by [@hengl_global_2018]. `terra` [@terra] was used  to derive the covariates slope, aspect, and latitude from this elevation data.

The covariate layers temperature and precipitation were based on WorldClim v2 [@fick_worldclim_2017] and retrieved via `geodata` [@geodata]. WorldClim provides long-term monthly averages of climatic data at a 1 km resolution on a global scale, from which mean annual temperature (ºC) and total annual precipitation (mm) were derived as covariates.

```{r}
#| label: load-covariate-data-1

# load DEM raster
dem <- rast('data/covariates/dtm_elevation_merit.dem_m_250m_s0..0cm_2017_v1.0.tif') %>%
  crop(germany_4326)

# calculate slope 
slope <- terrain(dem, v = "slope", unit = "degrees")
slope <-  project(slope, crs, method='bilinear') 

# calculate aspect
aspect <- terrain(dem, v = "aspect", unit = "degrees")

# latitude
latitude <- rast(list(latitude = init(dem, "y")))

# WorldClim (temperature and precipitation)
temp <- worldclim_country(country='germany', var='tavg', path='data/covariates/worldclim', res=0.5)
precip <- worldclim_country(country='germany', var='prec', path='data/covariates/worldclim', res=0.5)

temp_annual <- mean(temp) # mean annual temperature
precip_annual <- sum(precip) # total annual precipitation
```

Land cover dissimilarity was based on data by the Copernicus Global Land Service [@buchhorn_copernicus_2020], which contains a global land cover classification for 2015 at a 100 m resolution. To reduce computational demand, it was reprojected to the target resolution of 250 m via mode resampling. Based on this raster, the statistic "dissimilarity" from image texture measures from grey-level co-occurrence matrices (GLCM) was computed. Dissimilarity was calculated across all directions using a 9x9 window, to capture land cover heterogeneity in a radius of approximately 1 km$^2$.

```{r}
#| label: lc-and-dissimilarity

# set option to redo the dissimilarity calculation (takes a few minutes) or load existing raster
redo_lc_dissim <- FALSE

# set path
lc_dissim_path <- "data/covariates/lc_dissim.tif"

# load land cover data and crop to Germany
land_cover <- rast('data/covariates/PROBAV_LC100_global_v3.0.1_2015-base_Discrete-Classification-map_EPSG-4326.tif') %>%
  crop(germany_4326)

# obtain LC dissimilarity
# check if file exists AND redo flag is FALSE --> just load existing raster
if (file.exists(lc_dissim_path) && !redo_lc_dissim) {
  
  lc_dissim <- rast(lc_dissim_path)
  
} else {
  
  # reproject LC to target resolution (100 -> 250 m) to speed up processing
  land_cover_250m <- project(land_cover, slope, method='mode') # via majority voting

  # transform to RasterLayer (from raster package) - required for glcm
  land_cover_rast <- raster(land_cover_250m)
  
  # dissimilarity
  dissim_rast <- glcm(
    land_cover_rast,
    window = c(9,9), # take the 9x9 pixels forming the square around current pixel -> dissim in 1.125 km2 radius)
    shift = list(c(1,0), c(0,1), c(1,1), c(-1,1)), # average 4 directions (up, down, left, right)
    statistics = "dissimilarity"
    )
  
  # convert back to terra
  lc_dissim <- rast(dissim_rast)
  
  # save
  writeRaster(lc_dissim, lc_dissim_path, overwrite=TRUE)
}

```

Rasters of clay and silt content (g kg$^{-1}$), pH, and soil water content (%) at field capacity were obtained at a 250 m resolution from SoilGrids (https://soilgrids.org/). To approximate the sampling depth of LUCAS measurements (20 cm), weighted averages of the 0-5 and 5-15 cm layers were calculated for the four variables.

The soil respiration rate in g C m$^{-2}$ y$^{-1}$ at a 1 km resolution was obtained from @earth_science_data_systems_global_2025.

For the human footprint, a raster of the 2015 human footprint at a 1 km resolution developed by @mu_global_2022 was used.

The Normalized Difference Vegetation Index (NDVI) was based on MODIS Terra Vegetation Indices [@didan_modisterra_2021], which provides NDVI data at a 250 m resolution every 16 days. The NDVI value represents the 3-year growing season (May-September) mean from 2013 to 2015, in order to capture the long-term vegetation trends. This mean was calculated and the resulting raster retrieved via Google Earth Engine.

```{r}
#| label:  load-covariate-data-2

# for next covariates, take weighted average of 0-5 cm and 5-15 cm raster
average_rasters <- function(rast1_path, rast2_path, weight1 = 1/3, weight2 = 2/3, resampling='bilinear') { 
  
  # load and crop first
  rast1 <- rast(rast1_path) %>%
    crop(germany_4326)
    
  # load and crop second, make sure to match grid of first via resampling
  rast2 <- rast(rast2_path) %>%
    crop(germany_4326) %>%
    project(rast1, method='bilinear')
  
  averaged_raster <- weight1 * rast1 + weight2 * rast2
  
  return(averaged_raster)
}

clay <- average_rasters('data/covariates/clay_0-5.tif', 'data/covariates/clay_5-15.tif')
silt <- average_rasters('data/covariates/silt_0-5.tif', 'data/covariates/silt_5-15.tif') 
ph <- average_rasters('data/covariates/ph_0-5.tif', 'data/covariates/ph_5-15.tif') / 10
# soil water content (SWC)
swc <- average_rasters('data/covariates/swc_0-5.tif', 'data/covariates/swc_5-15.tif') / 10

# soil respiration
s_resp <- rast('data/covariates/soil_Rs_mean.tif') %>%
  crop(germany_4326)

# human footprint (HFP)
hfp <- rast('data/covariates/hfp2015.tif') 
hfp <- crop(hfp, st_transform(germany, crs(hfp)))

# Normalized Difference Vegetation Index (NDVI)
ndvi <- rast('data/covariates/NDVI_Germany_2013_2015.tif')

```

Once the raster data was obtained for all predictors, all layers were projected to the same CRS (EPSG 25832, UTM 32N) and resolution (250 m) using bilinear resampling. The reference raster chosen for this alignment was the slope since it already had the target resolution of 250 m.

To account for differences in land cover, four natural land cover types (from the land cover data by @buchhorn_copernicus_2020) were encoded as dummy variables: cropland, forest, herbaceous vegetation (natural grassland), and wetland.

```{r}
#| label: prepare-raster-stack

# same CRS and aligned

# set slope raster as template to align all other rasters with (because it has target resolution 250m)
slope <-  project(slope, crs, method='bilinear') 

# add to stack
predictor_stack <- c(slope)

# realign all remaining raster layers
predictors <- list(dem, aspect, temp_annual, precip_annual, latitude, lc_dissim, clay, silt, ph, swc, s_resp, hfp, ndvi)
for (predictor in predictors) {
  predictor <- project(predictor, slope, method="bilinear")
  predictor_stack <- append(predictor_stack, predictor)
}

# reproject land cover (categorical, therefore use different reprojection method)
land_cover_reproj <- project(land_cover, slope, method='mode') # majority voting

# add dummy variables encoding major land cover types
forest_codes <- c(111,112,113,114,115,116,121,122,123,124,125,126)
forest <- land_cover_reproj %in% forest_codes  # returns binary raster (True/False)
grassland <- land_cover_reproj == 30
wetland <- land_cover_reproj == 90
cropland <- land_cover_reproj == 40

# add to predictor stack (multiply by one to make True to 1 and False to 0)
predictor_stack <- c(predictor_stack, forest*1, grassland*1, wetland*1, cropland*1)

# adjust names
names(predictor_stack) <- c('slope', 'elevation', 'aspect', 'mean_temp', 'total_precip', 'latitude', 'lc_dissim', 'clay', 'silt', 'ph', 'swc', 's_resp', 'hfp', 'ndvi', 'forest', 'grassland', 'wetland', 'cropland')

# mask with Germany (extract exact shape)
predictor_stack <- mask(predictor_stack, germany)

```

For a visualization of predictor ranges and (spatial) distributions, please refer to @fig-plot-predictor-stack in the appendix.

## Models

In this study, we compared the performance of random forest (RF) and cubist machine learning for soil C:N prediction. For this purpose, predictor values were extracted for each LUCAS sampling point. Models were then trained to predict the C:N ratio from these predictors.

```{r}
#| label: prepare-model-data

# extract data for each point from raster stack
predictor_values <- terra::extract(predictor_stack, lucas)[,-1] # remove ID

# combine with C:N data and remove unnecessary colums
data <- cbind(lucas['cn_ratio'], predictor_values) %>%
  drop_na() %>% 
  st_drop_geometry() 

```

K-fold cross-validation (CV) was used for hyperparameter tuning and estimating accuracy metrics. In k-fold CV, the dataset is split into k approximately equal parts ("folds"), and a model is iteratively fit on k-1 folds while the remaining fold is used for validation. The model accuracy metrics are then obtained by averaging the accuracy metrics for predictions (of the validation fold samples) across the folds. This CV approach offers a more robust estimate of model accuracy than a single training and testing split.

```{r}
#| label: model-cross-validation

# set seeds for reproducibility (one for each model run)
set.seed(123)
seeds <- vector(mode = "list", length = 11)
for(i in 1:10) seeds[[i]] <- sample.int(1000, 5)
seeds[[11]] <- sample.int(1000, 1)

# 10-fold CV (split model into 10 folds, run 10 times, each time, 1 fold is test while remaining 9 are used for training)
train_control <- trainControl(method = "cv", 
                              number = 10,
                              seeds = seeds,
                              savePredictions = "final"
                              )

# train random forest with 10-fold CV
rfModel <- train(cn_ratio ~ ., # C:N ratio as a function of all predictors
                 data = data, 
                 method = "rf", # random forest
                 tuneLength = 5,
                 ntree = 500,
                 trControl = train_control, # implement CV as defined above
                 importance = TRUE) # compute permutation importance


# same for cubist
set.seed(123)
seeds <- vector(mode = "list", length = 11)
for(i in 1:10) seeds[[i]] <- sample.int(1000, 9)
seeds[[11]] <- sample.int(1000, 1)

train_control <- trainControl(method = "cv", 
                              number = 10,
                              seeds = seeds,
                              savePredictions = "final"
                              )

cubModel <- train(cn_ratio ~ ., # C:N ratio as a function of all predictors
                 data = data, 
                 method = "cubist", 
                 trControl = train_control, 
                 importance = TRUE) 

```

This k-fold CV was implemented using the `caret` [@caret] package in R, setting k to 10. Caret automatically conducted hyperparameter tuning, and for the parameter `mtry` in random forest, i.e., the number of predictors used at each split, the lowest Root Mean Squared Error (RMSE) was found for `mtry` = `r rfModel$bestTune$mtry`. The number of trees was set to 500. For the cubist model, hyperparameter tuning showed that `r cubModel$bestTune$committees` committees and `r cubModel$bestTune$neighbors` neighbors led to the lowest RMSE. 

After the 10-fold CV, a random forest and a cubist model with these optimal hyperparameter values was trained on the full dataset to obtain the relative variable importance scores.

## Model performance

Following @khosravani_soil_2026, the model accuracy was evaluated based on four metrics:

-   **coefficient of determination (R$^2$)**, which indicates how much of the variance can be explained by the model
-   **root mean squared error (RMSE)**, which measures the prediction error in units of the predicted variable, penalizing larger errors
-   **Bias**, which shows whether the model systematically over- or underestimates the target variable
-   **concordance correlation coefficient (CCC)**, which quantifies the agreement between observed and predicted values

Following the approach of @zhou_prediction_2021, accuracy metrics were calculated as the mean from the 10-fold cross-validation predictions to have more robust estimates of model accuracy.

```{r}
#| label: calculate-accuracy-measures

# extract observations, predictions, and fold columns from best-fit model
rf_preds <- rfModel$pred %>%
  select(pred, obs, Resample)

# calculate accuracy metrics per fold
rf_preds_accuracy <- rf_preds %>%
  group_by(Resample) %>%
  summarise(
    RMSE = sqrt(mean((obs - pred)^2)),
    R2 = 1 - ((mean((obs - pred)^2))/(mean((obs - mean(rf_preds$obs))^2))),
    Bias = mean(obs - pred),
    CCC = (2 * cor(rf_preds$obs, rf_preds$pred) * sd(rf_preds$obs) * sd(rf_preds$pred)) /
  (var(rf_preds$obs) + var(rf_preds$pred) + (mean(obs) - mean(pred))^2)
) %>% 
  arrange(Resample) # make sure it is ordered by fold

# same for cubist
cubist_preds <- cubModel$pred %>%
  select(pred, obs, Resample)

cubist_preds_accuracy <- cubist_preds %>%
  group_by(Resample) %>%
  summarise(
    RMSE = sqrt(mean((obs - pred)^2)),
    R2 = 1 - ((mean((obs - pred)^2))/(mean((obs - mean(cubist_preds$obs))^2))),
    Bias = mean(obs - pred),
    CCC = (2 * cor(cubist_preds$obs, cubist_preds$pred) * sd(cubist_preds$obs) * sd(cubist_preds$pred)) /
  (var(cubist_preds$obs) + var(cubist_preds$pred) + (mean(obs) - mean(pred))^2)
) %>% 
  arrange(Resample) 

# combine in one df
rf_preds_accuracy$Model <- "random forest"
cubist_preds_accuracy$Model <- "cubist"
accuracies <- rbind(rf_preds_accuracy, cubist_preds_accuracy)

# average within each model across folds
accuracies_summarized <- accuracies %>%
  group_by(Model) %>%
  # summarise(across(everything(), mean)) %>%
  summarise(
    across(where(is.numeric), mean, .names = "{.col}_mean"),
    across(where(is.numeric), sd,   .names = "{.col}_sd")
  ) %>%
  # summarize in one column with ±
  transmute(
    Model = Model,
    RMSE = glue("{round(RMSE_mean, 2)} ± {round(RMSE_sd, 2)}"),
    R2 = glue("{round(R2_mean, 3)} ± {round(R2_sd, 2)}"),
    Bias = glue("{round(Bias_mean, 2)} ± {round(Bias_sd, 2)}"),
    CCC = glue("{round(CCC_mean, 2)} ± {round(CCC_sd, 3)}")
  )
  
```

```{r}
#| label: paired-t-tests

# for R2
shapiro_r2 <- shapiro.test(rf_preds_accuracy$R2 - cubist_preds_accuracy$R2)
t_test_r2 <- t.test(rf_preds_accuracy$R2, cubist_preds_accuracy$R2, paired = TRUE)


# for RMSE
shapiro_rmse <- shapiro.test(rf_preds_accuracy$RMSE - cubist_preds_accuracy$RMSE)
t_test_rmse <- t.test(rf_preds_accuracy$RMSE, cubist_preds_accuracy$RMSE, paired = TRUE)

# for Bias
shapiro_bias <- shapiro.test(rf_preds_accuracy$Bias - cubist_preds_accuracy$Bias)
t_test_bias <-t.test(rf_preds_accuracy$Bias, cubist_preds_accuracy$Bias, paired = TRUE)

# for CCC
shapiro_ccc <- shapiro.test(rf_preds_accuracy$CCC - cubist_preds_accuracy$CCC)
t_test_ccc <-t.test(rf_preds_accuracy$CCC, cubist_preds_accuracy$CCC, paired = TRUE)

```

To evaluate whether the models' performance was significantly different, paired t-tests were conducted on the accuracy metric distributions across the CV folds. A Shapiro-Wilk test was used beforehand to ensure an approximately normal distribution of differences between the paired metrics (p = `r round(shapiro_r2$p.value,3)` for R$^2$, p = `r round(shapiro_rmse$p.value,3)` for RMSE, p = `r round(shapiro_bias$p.value,3)` for Bias, p = `r round(shapiro_ccc$p.value,3)` for CCC). However, it needs to be acknowledged that conducting a paired t-test on the values from 10-fold CV might have an increased risk of Type 1 error (suggesting a significant difference when there is none), because while the test data the metrics are based on is independent, the training data is not completely since it partially overlaps [@dietterich_approximate_1998].

## Mapping

To obtain mean and uncertainty maps, we used a bootstrapping approach as done by @khosravani_soil_2026. Due to computational constraints, 30 iterations were used for this process. In each iteration, a cubist and a random forest model were trained on a bootstrapped sample of the data (representing on average 63.2% of the complete data) and then used to make spatial predictions based on the predictor stack. From the resulting 30 maps per model, the mean predicted C:N as well as an uncertainty map based on the standard deviation were calculated. To investigate possible spatial patterns between predictions from the two models, a difference map was computed by substracting the mean cubist from the mean RF map.

```{r}
#| label: bootstrapping

# TAKES LONG (around 4 hours)! only redo if something about data changed - otherwise files will be loaded from disk (if available)
redo_bootstrap <- FALSE

# set number of bootstrapping runs
b_max = 30

n = nrow(data)

if (redo_bootstrap) {
  
  # downsample raster stack for testing
  # predictor_stack_coarse <- terra::aggregate(predictor_stack, fact = 3)

  for (b in seq(1,b_max)) {
    
    # progress report
    cat("Bootstrap iteration:", b, "of", b_max, "\n")
    
    # get indexes via bootstrap sampling (with replacement)
    idxs <- sample(n, size=n, replace = TRUE)
    
    # subset data
    boot_data <- data[idxs,]
    
    # train models (with best hyperparameter tuning values) on it
    rf_model <- randomForest(
      cn_ratio ~ .,      # C:N as function of all other columns (predictors)
      data = boot_data,
      mtry = rfModel$bestTune$mtry,
      ntree = 500       # number of trees
      )
    cubist_model <- cubist(
      x = select(boot_data, !cn_ratio), 
      y = boot_data$cn_ratio,
      committees = cubModel$bestTune$committees
      )
    
    # spatial prediction
    map_rf <- terra::predict(
      object = predictor_stack, # what to predict from
      model  = rf_model,
      cores=4, # parallel processing
      cpkgs = "randomForest", # package needed to run prediction
      filename = paste0('data/bootstrap_prediction_layers/rf_', b, '.tif' ), # write to disk instead of keeping in memory
      overwrite = TRUE
      )
    map_cubist <- terra::predict(
      object = predictor_stack, 
      model  = cubist_model,
      cores=4,
      cpkgs = "cubist", 
      filename = paste0('data/bootstrap_prediction_layers/cub_', b, '.tif' ), 
      overwrite = TRUE
    )
  }
  
  # load bootstrapping map predictions
  path <- 'data/bootstrap_prediction_layers/'
  rf_boot_stack <- rast(paste0(path, list.files(path, pattern='rf*')))
  cub_boot_stack <- rast(paste0(path, list.files(path, pattern='cub*')))
  
  # mask cubist stack with Germany (because it also predicts for NA values)
  cub_boot_stack <- mask(cub_boot_stack, germany)
  
  # compute mean and standard deviation across layers
  rf_mean_map <- app(rf_boot_stack, mean)
  rf_sd_map   <- app(rf_boot_stack, sd)
  
  cub_mean_map <- app(cub_boot_stack, mean)
  cub_sd_map   <- app(cub_boot_stack, sd)
  
  # difference map
  diff_map <- rf_mean_map - cub_mean_map
  diff_map_abs <- abs(diff_map)
  
  # save files
  writeRaster(rf_mean_map, 'data/bootstrap_summaries/rf_mean_map.tif', overwrite = TRUE)
  writeRaster(rf_sd_map, 'data/bootstrap_summaries/rf_sd_map.tif', overwrite = TRUE)
  writeRaster(cub_mean_map, 'data/bootstrap_summaries/cub_mean_map.tif', overwrite = TRUE)
  writeRaster(cub_sd_map, 'data/bootstrap_summaries/cub_sd_map.tif', overwrite = TRUE)
  writeRaster(diff_map, 'data/bootstrap_summaries/diff_map.tif', overwrite = TRUE)
  writeRaster(diff_map_abs, 'data/bootstrap_summaries/diff_map_abs.tif', overwrite = TRUE)
}

# load from disk
rf_mean_map <- rast('data/bootstrap_summaries/rf_mean_map.tif')
rf_sd_map   <- rast('data/bootstrap_summaries/rf_sd_map.tif')
cub_mean_map <- rast('data/bootstrap_summaries/cub_mean_map.tif')
cub_sd_map   <- rast('data/bootstrap_summaries/cub_sd_map.tif')
diff_map <- rast('data/bootstrap_summaries/diff_map.tif')
diff_map_abs <- rast('data/bootstrap_summaries/diff_map_abs.tif')

```


## Ecological plausibility

To illustrate a possible use case for C:N maps and evaluate the ecological plausibility of the spatial predictions, we tested whether predicted C:N ratios in grasslands were significantly higher if they were located in protected areas. For this purpose, a shapefile of Natura 2000 areas -- the major network of protected areas in Europe -- was obtained from @natura, which was rasterized using the reference slope grid. A separate grassland map from @grassland was used since the land cover used for the predictor stack did not include it as a distinct class. This raster was reprojected from 100 to 250 m via mode resampling. Then, the mean maps from both models were masked to obtain grassland C:N values in and outside protected areas.

```{r}
#| label: prepare-natura-grassland-df

# load Natura 2000 shapefile
natura_vect <- vect('data/SHP/Natura2000_end2024_epsg3035.shp') %>%
  project(crs) %>%
  crop(germany)

# rasterize
natura <- rasterize(natura_vect, slope, background=0, field=1)

# load grassland raster
grassland_eu <- rast('data/GRA_2015_100m_eu_03035_V1_2.tif')

# crop
grassland_ger <- grassland_eu %>%
  crop(st_transform(germany, crs(grassland_eu)))

# make binary raster (1 where grassland, 0 all others)
grassland_ger <- grassland_ger == 1

# reproject to match raster stack grid (250 m) using mode resampling --> only classify as grassland if majority of 100 pixels are grassland 
grassland_ger <- project(grassland_ger, slope, method='mode')


# function to mask rasters by grassland and natura, and convert to dataframe
# function developed by Claude Sonnet 4.6
create_cn_boxplot_df <- function(cn_raster1, cn_raster2,
                                  grassland_raster, natura_raster,
                                  model_names = c("model1", "model2")) {
  
  # Stack all rasters and extract values in one pass
  stack <- c(cn_raster1, cn_raster2, grassland_raster, natura_raster)
  vals  <- terra::values(stack, dataframe = TRUE, na.rm = FALSE)
  colnames(vals) <- c("cn1", "cn2", "grassland", "natura")
  
  # Build a valid-cell mask (exclude cells where any layer is NA)
  valid <- complete.cases(vals)
  vals  <- vals[valid, ]
  
  # Pivot the two CN columns into long format
  df <- bind_rows(
    data.frame(
      grassland = as.logical(vals$grassland),
      protected = as.logical(vals$natura),
      model     = model_names[1],
      cn_ratio  = vals$cn1
    ),
    data.frame(
      grassland = as.logical(vals$grassland),
      protected = as.logical(vals$natura),
      model     = model_names[2],
      cn_ratio  = vals$cn2
    )
  )
  
  df$model <- as.character(df$model)
  return(df)
}

# create
grass_natura_df <- create_cn_boxplot_df(
  cn_raster1       = rf_mean_map,
  cn_raster2       = cub_mean_map,
  grassland_raster = grassland_ger,
  natura_raster    = natura,
  model_names      = c("RF", "cubist")
) %>%
  filter(grassland == TRUE)
```

In order to compare the differences statistically, a Mann-Whitney U test was conducted to test for significance, and Cliff's Delta was calculated to obtain the effect size. For the latter, a random subset of 10,000 protected and unprotected pixels each was used due to computational limitations.

```{r}
#| label: stat-tests-grassland-natura


### for random forest

# filter
grass_natura_df_rf <- filter(grass_natura_df, grass_natura_df$model == 'RF')

# Mann–Whitney U
mwu_rf <- wilcox.test(cn_ratio ~ protected, data = grass_natura_df_rf)

# get effect size (on random sample)
set.seed(123)
sample_df <- grass_natura_df_rf %>%
  dplyr::group_by(protected) %>%
  dplyr::slice_sample(n = 10000)
cliff_rf <- cliff.delta(cn_ratio ~ protected, data = sample_df)

### same for cubist

# filter
grass_natura_df_cub <- filter(grass_natura_df, grass_natura_df$model == 'cubist')

# Mann–Whitney U
mwu_cub <- wilcox.test(cn_ratio ~ protected, data = grass_natura_df_cub)

# get effect size
sample_df <- grass_natura_df_cub %>%
  dplyr::group_by(protected) %>%
  dplyr::slice_sample(n = 10000)
cliff_cub <- cliff.delta(cn_ratio ~ protected, data = sample_df)
```

{{< pagebreak >}}

# Results

## Model performance

Random forest and cubist showed similar accuracy in the 10-fold cross-validation (@tbl-acc-table). RF had a marginally lower error (RMSE) and captured slightly more of the variability (R$^2$) than the cubist model, but the differences were not statistically different (t(9) = `r round(t_test_rmse$statistic[['t']], 2)`, p = `r round(t_test_rmse$p.value, 2)` for RMSE; t(9) = `r round(t_test_r2$statistic[['t']], 2)`, p = `r round(t_test_r2$p.value, 2)` for R$^2$). 
The cubist model had a higher mean CCC value, however, which was also reflected in @fig-obs-vs-pred, where the highest density of points was closer to the 1:1 line for cubist. This difference was statistically significant (t(9) = `r round(t_test_ccc$statistic[['t']], 2)` p < 0.001).
A notable difference was found in the Bias, which showed that RF tended to overestimate C:N more often while cubist tended to underestimate it (see also @tbl-accuracy-by-fold in appendix), but both only to a small extent. This difference was also statistically significant (t(9) = `r round(t_test_bias$statistic[['t']], 2)` p = `r round(t_test_bias$p.value, 3)`).
Most metrics had low standard deviations, especially CCC, indicating stable model performance across folds, i.e., on different training and validation data. The highest relative standard deviation was found for Bias, which suggests that models trained on certain samples had a stronger systematic over- or underestimation.

```{r}
#| label: tbl-acc-table
#| tbl-cap: "Accuracy metrics of the two models. Values indicate mean and standard deviation from the 10-fold cross-validation."

knitr::kable(accuracies_summarized)
```

```{r}
#| label: fig-obs-vs-pred
#| fig-cap: "Observed vs. predicted values for random forest and cubist models. Predicted values are from out-of-box prediction during the 10-fold CV. Count refers to the number of sampling points represented by each plotted hexagon."


# two-panel plot with RF and cubist

# Bin size control + color palette
p1 <- ggplot(rf_preds, aes(obs, pred) ) +
  geom_hex(bins = 70) +
  scale_fill_continuous(limits=c(0,30), type = "viridis", name = 'count') +
  theme_bw() +
  xlab("observed") +
  ylab("predicted") + 
  ggtitle("random forest") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") #+
  # coord_equal() 

p2 <- ggplot(cubist_preds, aes(obs, pred) ) +
  geom_hex(bins = 70) +
  scale_fill_continuous(limits=c(0,30), type = "viridis", name = 'count') +
  theme_bw() +
  xlab("observed") +
  ylab("predicted") + 
  ggtitle("cubist") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") #+
  # coord_equal() 

(p1 + p2) + plot_layout(guides = "collect")

```


## Variable importance

```{r}
#| label: fig-variable-importance
#| fig-cap: "Relative importance of environmental covariates for random forest (RF) and cubist prediction of soil C:N. Derived from models trained on complete sample dataset."

# extract importance scores
var_imp_rf <- varImp(rfModel)$importance
var_imp_cub <- varImp(cubModel)$importance

# start new df with RF
imp_df <- tibble(
  variable = rownames(var_imp_rf),
  RF = var_imp_rf$Overall
) 
  
# add cubist scores
imp_df <- tibble(
  variable = rownames(var_imp_cub),
  cubist = var_imp_cub$Overall
  ) %>%
  full_join(imp_df, by = join_by(variable)) %>% # combine with RF scores
  pivot_longer( # modify format
    cols = c(RF, cubist),
    names_to = 'model',
    values_to = 'importance'
  ) %>%
  arrange(desc(model), desc(importance))

# make variable into ordered factor column based on RF importance scores (to order plot by descending importance scores)
rf_order <- imp_df %>%
  filter(model == "RF") %>%
  arrange(importance) %>%
  pull(variable)
imp_df$variable <- factor(imp_df$variable, levels = rf_order)

ggplot(imp_df, aes(variable, importance, fill=model)) +
  geom_col(position='dodge') +
  coord_flip() +
  xlab(element_blank()) +
  ylab('Relative importance [%]')
```

pH was by far the most important predictor in both models, while NDVI was ranked second (@fig-variable-importance). The next most important predictors for RF was the forest dummy, which had a lower rank for cubist. While not identical in their exact ranking, both models had many of the same variables with moderate importance (between 20 and 45%). Notably, soil water content (SWC), total annual precipitation, and silt content, which had a comparatively low ranking for RF, were among the most important variables for cubist. The least important numeric predictors in both models were soil respiration and aspect. The dummy variables for cropland, wetland, and grassland also ranked low.

## Spatial predictions

```{r}
#| label: fig-plot-mean-maps
#| fig-cap: "C:N ratio across Germany predicted by random forest and cubist machine learning. Values represent mean prediction from 30 bootstrapping iterations."

# determine shared value limits
zlim <- range(c(values(rf_mean_map), values(cub_mean_map)), na.rm = TRUE)

p1 <- ggplot() +
  geom_spatraster(data = rf_mean_map) +
  scale_fill_viridis_c(limits = zlim, 
                       na.value = "transparent",
                       name = "C:N") +
  ggtitle("random forest") +
  theme_void()

p2 <- ggplot() +
  geom_spatraster(data = cub_mean_map) +
  scale_fill_viridis_c(limits = zlim, 
                       na.value = "transparent",
                       name = "C:N") +
  ggtitle("cubist") +
  theme_void()

# combine plots with shared colorbar
(p1 | p2) + plot_layout(guides = "collect")
```

The mean predicted C:N ratios derived from bootstrapping showed similar spatial distributions for random forest and cubist prediction (@fig-plot-mean-maps). Both models suggested distinct large-scale patterns of areas with elevated and with low ratios. Higher values were mostly located across North-Eastern Germany. Lower values were found near coasts and in the South-East. Overall, random forest had a higher mean of predicted values (`r show_mean_sd(rf_mean_map)`) than cubist (`r show_mean_sd(cub_mean_map)`), and its predicted value range was higher (`r round(min(values(rf_mean_map), na.rm=TRUE),2)` - `r round(max(values(rf_mean_map), na.rm=TRUE),2)`) compared to cubist (`r round(min(values(cub_mean_map), na.rm=TRUE),2)` - `r round(max(values(cub_mean_map), na.rm=TRUE),2)`).

```{r}
#| label: fig-plot-sd-maps
#| fig-cap: "C:N ratio across Germany predicted by random forest and cubist machine learning. Values represent standard deviation of predictions from 30 bootstrapping iterations."


# Determine shared limits
zlim <- range(c(values(rf_sd_map), values(cub_sd_map)), na.rm = TRUE)

p1 <- ggplot() +
  geom_spatraster(data = rf_sd_map) +
  scale_fill_viridis_c(limits = zlim, 
                       na.value = "transparent",
                       name = "C:N\nstandard\ndeviation") +
  ggtitle("random forest") +
  theme_void()

p2 <- ggplot() +
  geom_spatraster(data = cub_sd_map) +
  scale_fill_viridis_c(limits = zlim, 
                       na.value = "transparent", 
                       name = "C:N\nstandard\ndeviation") +
  ggtitle("cubist") +
  theme_void()

# Combine with patchwork, collecting the shared legend
(p1 + p2) + plot_layout(guides = "collect")

```

With a mean standard deviation of `r show_mean_sd(rf_sd_map)`, random forest predictions were on average less uncertain than those from cubist, which had a mean standard deviation of `r show_mean_sd(cub_sd_map)`. Especially for the cubist prediction, the spatial pattern of uncertainty resembled the pattern of higher predicted C:N values (@fig-plot-sd-maps).Maximum standard deviations in the cubist map were located where the random forest did not make predictions due to missing data, mostly where large water bodies are located.

```{r}
#| label: fig-diff-map
#| fig-cap: "Difference map between random forest and cubist predictions. Blue values indicate higher random forest predictions while red values indicate higher cubist predictions. Computed from the mean maps from 30 bootstrapping iterations."

diff_df <- as.data.frame(diff_map, xy = TRUE)

# plot
ggplot(diff_df) +
  geom_spatraster(data = diff_map) +
  scale_fill_gradient2(
    low = "#b2182b",
    mid = "white",
    high = "#2166ac",
    midpoint = 0,
    limits = c(-4, 4),
    na.value = "transparent",
    name = "Prediction\ndifference"
  ) +
  theme_void()

# get percentage of certain difference ranges
vals <- values(diff_map)
total <- sum(!is.na(vals))
fraction_small_diff <- sum(vals >= -1 & vals <= 1, na.rm = TRUE) / total
fraction_very_small_diff <- sum(vals >= -0.5 & vals <= 0.5, na.rm = TRUE) / total
fraction_medium_diff <- sum(vals >= -2 & vals <= 2, na.rm = TRUE) / total
fraction_rf_higher <- sum(vals > 0, na.rm = TRUE) / total
fraction_cub_higher <- sum(vals < 0, na.rm = TRUE) / total

```
 
The mean difference in predictions between the two models was `r show_mean_sd(diff_map)`, and ranged from `r round(min(values(diff_map), na.rm = TRUE), 2)` to `r round(max(values(diff_map), na.rm = TRUE), 2)`. Most pixels (`r round(fraction_very_small_diff * 100, 2)`%) had only small absolute differences of up to 0.5. Overall, random forest had higher predictions in a majority of pixels (`r round(fraction_rf_higher * 100, 2)`%), see @fig-diff-map. We found a moderate pixelwise correlation between the absolute difference between models and the uncertainty (r = `r round(cor(values(diff_map_abs), values(rf_sd_map), use = "na.or.complete")[1,1], 2)` for random forest, r = `r round(cor(values(diff_map_abs), values(cub_sd_map), use = "na.or.complete")[1,1], 2)` for cubist).

```{r}
#| label: fig-plot-by-land-cover
#| fig-cap: "Distribution of mean and standard deviation (uncertainty) of predicted C:N by land cover class. Values represent mean/standard deviation from 30 bootstrapping iterations."

# make data frame
df <- tibble(
  grassland = values(grassland)[,1],
  forest = values(forest)[,1],
  cropland = values(cropland)[,1],
  wetland = values(wetland)[,1],
  rf_mean = values(rf_mean_map)[,1],
  rf_sd = values(rf_sd_map)[,1],
  cub_mean = values(cub_mean_map)[,1],
  cub_sd = values(cub_sd_map)[,1]
)

df <- df %>%
  mutate(
    landcover = case_when(
      grassland == 1 ~ "Grassland",
      forest == 1 ~ "Forest",
      cropland == 1 ~ "Cropland",
      wetland == 1 ~ "Wetland",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(landcover))

# reorganize df for easier plotting
df_plot <- bind_rows(
  df %>% transmute(landcover, value = rf_mean, model = "RF", type = "Mean"),
  df %>% transmute(landcover, value = cub_mean, model = "cubist", type = "Mean"),
  df %>% transmute(landcover, value = rf_sd, model = "RF", type = "Uncertainty"),
  df %>% transmute(landcover, value = cub_sd, model = "cubist", type = "Uncertainty")
)

# plot
ggplot(df_plot, aes(x = landcover, y = value, fill = model)) +
  geom_boxplot(outlier.size = 0.3) +
  facet_wrap(~type, scales = "free_y", ncol=1) +
  labs(
    x = "Land cover type",
    y = "C:N",
    fill = "Model"
  ) +
  theme_minimal()

```

Both models mostly captured the same ranking of land cover types in terms of mean prediction and uncertainty (@fig-plot-by-land-cover). Highest mean values were found in forests, which also had the highest uncertainty. Wetlands had a comparable level of uncertainty, as well as the second highest mean C:N. Third ranked in both aspects were grasslands and lastly cropland. The boxplot highlighted again how random forest overall had a lower uncertainty, with cubist values often exhibiting larger ranges.

## Ecological plausibility

```{r}
#| label: fig-natura-boxplot
#| fig-cap: "Distribution of predicted values in grasslands within and outside Natura 2000 protected areas. Values at the top show mean ± standard deviation. C:N values refer to the mean derived from 30 bootstrapping iterations."

# compute mean and sd per group
summary_df <- grass_natura_df %>% 
  group_by(model, protected) %>%
  summarise(
    label = glue("{round(mean(cn_ratio), 2)} ± {round(sd(cn_ratio), 2)}"),
    mean = mean(cn_ratio),
    y = max(cn_ratio) * 1.05,   # position above box
    .groups = "drop"
  )

# get percentage table
p <- prop.table(table(grass_natura_df$protected)) * 100

# plot
ggplot(grass_natura_df, aes(protected, cn_ratio, fill = model)) +
  geom_boxplot(position = position_dodge(width = 0.85)) +
  ylab('C:N') +
  geom_text(
    data = summary_df,
    aes(label = label, y = y, group = model),
    position = position_dodge(width = 0.85),
    size = 3
  )

```

Both models predicted slightly higher C:N ratios in Natura 2000 protected (`r as.numeric(round(p["TRUE"],2))`%) grasslands than in unprotected (`r as.numeric(round(p["FALSE"],2))`%) grasslands (@fig-natura-boxplot). The difference in means was slightly higher for RF (`r round(summary_df$mean[4] - summary_df$mean[3],2)`) than for cubist (`r round(summary_df$mean[2] - summary_df$mean[1],2)`). The differences were found to be significant in both models (Mann–Whitney U, p < 0.01), but with a small effect size for random forest (Cliff's Delta `r round(cliff_rf$estimate, 3)`) and negligible effect size for cubist (Cliff's Delta `r round(cliff_cub$estimate, 3)`).

{{< pagebreak >}}

# Discussion

## Model performance

The predictive performance achieved in this study (random forest: RMSE = `r accuracies_summarized$RMSE[accuracies_summarized$Model == "random forest"]`, R² = `r accuracies_summarized$R2[accuracies_summarized$Model == "random forest"]`, CCC = `r accuracies_summarized$CCC[accuracies_summarized$Model == "random forest"]`, Bias = `r accuracies_summarized$Bias[accuracies_summarized$Model == "random forest"]`; cubist: RMSE = `r accuracies_summarized$RMSE[accuracies_summarized$Model == "cubist"]`, R$^2$ = `r accuracies_summarized$R2[accuracies_summarized$Model == "cubist"]`, CCC = `r accuracies_summarized$CCC[accuracies_summarized$Model == "cubist"]`, Bias = `r accuracies_summarized$Bias[accuracies_summarized$Model == "cubist"]`) is closely comparable to the test metrics reported by @khosravani_soil_2026, who achieved R² = 0.63, CCC = 0.76, RMSE = 2.55 and Bias = −0.02 for C:N prediction across Germany using a cubist-based approach combined with bootstrapping. This close agreement suggests that similar levels of predictive performance can be achieved across different modeling approaches when applied to comparable datasets and spatial scales.

Despite this similarity, the evaluation metrics reveal subtle but consistent differences in model behaviour. The slightly lower RMSE and higher R² of random forest indicate a marginally better global fit, whereas cubist shows a slightly higher concordance with observed values (CCC). These differences reflect known characteristics of the algorithms: random forest stabilises predictions through ensemble averaging, while cubist combines rule-based partitioning with local linear models, which may better preserve local relationships [@khaledian_selecting_2020; @wadoux_machine_2020].

Bias patterns further highlight systematic differences between the models. Random forest exhibited a slight negative bias (`r accuracies_summarized$Bias[accuracies_summarized$Model == "random forest"]`), indicating a tendency to overestimate C:N values, whereas cubist showed a slight positive bias (`r accuracies_summarized$Bias[accuracies_summarized$Model == "cubist"]`), suggesting a tendency towards underestimation. Although small in magnitude, these deviations were statistically significant and indicate consistent differences in how both models represent the distribution of C:N values. This is particularly relevant for ratio-based variables, where small deviations in the prediction of carbon or nitrogen can propagate into the resulting ratio.

In contrast, @ballabio_mapping_2019 reported substantially higher predictive performance, with an R² of 0.91 for the C:N ratio at the European scale based on LUCAS data. Such an outstanding level of model fit is likely related to the use of Gaussian process regression in combination with a comprehensive set of environmental covariates, including remote sensing predictors, which provide spatially continuous information on environmental gradients and have been shown to support large-scale digital soil mapping [@hong_path_2026; @radocaj_open_2024]. However, these high values should be interpreted in light of differences in modeling framework, predictor composition and spatial support, all of which can strongly influence reported performance metrics.

More generally, the interpretation of performance metrics requires careful consideration. As emphasized by @wadoux_spatial_2021, cross-validation-based accuracy measures primarily reflect model performance under the given sampling design and may not provide an unbiased estimate of map accuracy across the full target domain. Nevertheless, when sampling locations are not strongly clustered and reasonably represent the environmental conditions of the study area, cross-validation estimates can still approximate true predictive accuracy relatively well. In this context, the reported RMSE, R², CCC and Bias values are most useful for the relative comparison between random forest and cubist, while a fully unbiased assessment of map accuracy would require an independent, probability-based validation dataset collected after model development, for example through stratified random sampling across land-use types and environmental gradients, as proposed by @wadoux_spatial_2021.

## Variable importance

Across both modeling approaches, soil pH and NDVI emerged as the dominant predictors, with identical ranking positions in random forest and cubist. In cubist, pH accounted for the highest relative importance (100%), whereas NDVI contributed more strongly in random forest. The third-ranked numerical predictor differed between models was soil water content in cubist and latitude in random forest.

The prominence of soil pH aligns with @khosravani_soil_2026 and is consistent with its well-established role in regulating key biogeochemical processes relevant to soil stoichiometry. By contrast, the synthesis by @hong_path_2026 reported comparatively low importance of pH across studies, underscoring the context dependency of variable importance patterns.

The consistently high contribution of NDVI in the present study represents a key result. As a proxy for vegetation productivity and biomass dynamics, NDVI links aboveground processes to soil organic matter inputs. Its lower importance in @khosravani_soil_2026 may be attributable to differences in predictor construction, as the multi-year growing-season median applied here likely reduces short-term variability and enhances the representation of stable vegetation–soil interactions. This interpretation is consistent with @pouladi_mapping_2019, where NDVI was included among the most frequently used and relevant remote-sensing predictors in cubist- and random-forest-based soil mapping, forming part of the key environmental covariates across models. It is further supported by @wang_mapping_2025, who demonstrated the strong predictive contribution of multi-temporal optical and radar data for soil C:N mapping.

Despite the central role of microbial processes in soil C:N dynamics [@zechmeister-boltenstern_application_2015], soil respiration exhibited low importance in both the present study and the modeling framework of @khosravani_soil_2026, with substantially lower contributions in cubist than in random forest. This pattern likely reflects the high temporal variability and environmental sensitivity of respiration-related microbial processes, which limits their usefulness as stable spatial predictors [@davidson_temperature_2006; @chernov_dynamics_2020]. In addition, the stronger contribution in random forest may indicate its greater capacity to capture weak, non-linear and interaction-driven signals compared to the more structure-constrained cubist approach.

Land-cover effects differed between models: forest and cropland were relevant in both, whereas grassland and wetland contributed only in random forest. This pattern likely reflects both predictor representation and model structure. While categorical variables capture discrete classes, land-cover dissimilarity metrics used by @khosravani_soil_2026 represent spatial heterogeneity more continuously. In addition, variable-importance measures are model-specific and sensitive to predictor correlation [@strobl_bias_2007; @strobl_conditional_2008; @wadoux_machine_2020], limiting direct comparability between algorithms.

Taken together, these findings indicate that predictor importance does not follow a universal hierarchy but reflects the interaction of environmental conditions, covariate design and modeling approach. Accordingly, variable-importance metrics should be interpreted as indicators of predictive relevance rather than direct measures of process importance.

## Spatial patterns

The spatial predictions of soil C:N ratios revealed consistent large-scale gradients across Germany, with both models capturing patterns that broadly align with established controls of soil stoichiometry. Predicted values ranged from `r round(min(values(rf_mean_map), na.rm=TRUE),2)` to `r round(max(values(rf_mean_map), na.rm=TRUE),2)` for random forest and from `r round(min(values(cub_mean_map), na.rm=TRUE),2)` to `r round(max(values(rf_mean_map), na.rm=TRUE),2)` for cubist, falling within the general magnitude reported in previous studies. In particular, @khosravani_soil_2026 reported comparable national-scale C:N ranges for Germany (cubist: ~8–20), with most values occurring within a moderate range typical for mineral soils and higher values associated with forested and organic-rich environments. This suggests that both approaches are capable of reproducing broad-scale C:N distributions when key environmental predictors are included [@hong_path_2026; @zhou_prediction_2021].

A central pattern emerging from our results is the systematic compression of the predicted C:N range. Both models produced a narrower range compared to the observed soil data, indicating a tendency to underestimate variability. This behaviour is consistent with previous findings that machine-learning approaches tend to smooth extreme values and regress predictions towards the mean [@hong_path_2026; @sakhaee_spatial_2024].

Despite very similar overall model performance, no clear superiority of either algorithm can be identified. However, clear spatial discrepancies between models were evident, indicating that comparable global performance metrics do not necessarily translate into similar spatial predictions. The structured disagreement likely reflects differences in algorithmic learning strategy: random forest is an ensemble tree-based method capable of capturing complex nonlinear relationships [@breiman_random_2001], whereas cubist combines rule-based partitioning with local linear regression models [@noi_comparison_2017].

Model disagreement was spatially widespread across Germany, indicating that differences between algorithms occurred throughout the study area. However, disagreement values were not evenly distributed but showed clear spatial clustering, with higher concentrations of Δ values in specific physiographic regions. These hotspots were particularly evident in forested and topographically complex landscapes, including parts of southern Germany such as the Bavarian Forest and Alpine forelands, as well as selected areas of the Central Uplands (e.g. Harz, Hunsrück) and northern lowlands (e.g. northeastern lake districts, northwestern Geest). In contrast, regions with lower predicted C:N values showed a more diffuse spatial pattern and less pronounced clustering of disagreement, for example in parts of the Swabian-Franconian Alb and the Oldenburg–East Frisian Geest.

This spatial structuring suggests that model disagreement is not random but related to environmental gradients, particularly in geomorphologically complex and forest-dominated systems. While direct comparisons based on disagreement maps are rare, this pattern is consistent with findings from digital soil mapping studies showing that prediction uncertainty and model sensitivity tend to increase in heterogeneous landscapes, under high covariate variability and limited sample support [@nozari_digital_2024; @wadoux_machine_2020]. The magnitude of disagreement (|Δ|) further highlights regions of increased model sensitivity and can be interpreted as an indicator of model-dependent predictions, particularly where different algorithms respond differently to complex environmental conditions.

This interpretation is supported by the spatial distribution of prediction uncertainty. Random forest predictions were on average less uncertain (mean SD: `r show_mean_sd(rf_sd_map)`) than cubist predictions (`r show_mean_sd(cub_sd_map)`), indicating higher overall model stability. This pattern was consistent across all land-cover types, with lower uncertainty values for random forest in grassland (0.37 vs. 0.74), forest (0.64 vs. 1.14), cropland (0.31 vs. 0.47) and wetland systems (0.63 vs. 1.40). Similar tendencies have been reported in previous DSM studies, where random forest often showed more stable predictions compared to cubist in mapping soil properties [@khaledian_selecting_2020; @nozari_digital_2024].

Across both models, uncertainty was lowest in croplands and highest in forest and wetland systems, reinforcing the link between environmental heterogeneity and predictive uncertainty. Moderate correlations between absolute model differences and uncertainty (RF: r = `r round(cor(values(diff_map_abs), values(rf_sd_map), use = "na.or.complete")[1,1], 2)`; cubist: r = `r round(cor(values(diff_map_abs), values(cub_sd_map), use = "na.or.complete")[1,1], 2)`) further suggest that disagreement and uncertainty are partly related.

This supports previous findings that predictive uncertainty in digital soil mapping is closely linked to the representativeness of the environmental feature space [@khosravani_soil_2026; @schmidinger_validation_2023]. In addition, extreme uncertainty values in the cubist predictions partly coincide with areas where random forest did not produce predictions due to missing data, highlighting differences in model robustness to data gaps.

A key limitation underlying both spatial patterns and uncertainty structures is the configuration of the LUCAS dataset, with a mean sampling distance of approximately 7–8 km, which constrains the representation of fine-scale heterogeneity, particularly in variable systems such as grasslands and peat-influenced environments [@hong_path_2026; @pouladi_mapping_2019].

Overall, the results indicate that uncertainty should not be treated merely as a residual modeling artifact but as an informative component of spatial prediction. Areas of high uncertainty and strong model disagreement can serve as priority targets for improved sampling and model refinement, particularly in environmentally complex regions where current predictions are most constrained.

## Ecological plausibility

Predicted C:N ratios were on average higher in grasslands located within Natura 2000 areas compared to non-protected sites (Mann–Whitney U, p < 0.001; Cliff’s Delta = `r round(cliff_rf$estimate, 2)` for RF and `r round(cliff_cub$estimate, 2)` for cubist), indicating a systematic but small differentiation at the national scale. This supports the hypothesis that land-use intensity is reflected in soil stoichiometry, as protected grasslands are generally managed less intensively, with lower nitrogen inputs and disturbance. Accordingly, empirical studies show that intensively managed grasslands tend to exhibit narrower C:N ratios (~11–13), whereas more extensive systems display wider ratios (~13–16) due to lower nitrogen availability [@bluthgen_quantitative_2012; @herold_soil_2014]. Conservation management has likewise been linked to increased soil carbon storage and shifts in nutrient balances that can result in higher C:N ratios [@murindangabo_long-term_2026], consistent with ecological stoichiometry theory [@sardans_cnp_2012]. However, the small effect size indicates that protection status alone is a weak predictor of soil C:N at this scale. This likely reflects heterogeneous land-use intensity across both groups, combined with the uneven implementation of conservation measures within Natura 2000, which is often shaped by socio-economic and administrative constraints [@rouveyrol_agri-environmental_2024; @european_commission_directorate_general_for_agriculture_and_rural_development_evaluation_2020; @gorris_social-ecological_2025]. In addition, the absence of explicit management variables limits the representation of land-use effects. Moreover, predictive soil mapping approaches tend to underestimate the true range of C:N ratios, leading to compressed value distributions and reduced contrasts between land-use classes [@hong_path_2026], which likely contributes to the small observed effect size. Overall, the models capture ecologically plausible trends, but differentiation remains subtle due to both management heterogeneity and modeling-induced smoothing. Future studies must integrate spatially explicit management indicators (e.g. nitrogen inputs, mowing, grazing) and apply modeling approaches that reduce range compression, supported by targeted validation designs to isolate management effects.

{{< pagebreak >}}

# Conclusion

This study addressed key gaps in the spatial modeling of soil C:N ratios by providing a systematic comparison of random forest and cubist models at the national scale for Germany. The results show that both models achieved comparable predictive performance, indicating that similar levels of accuracy can be obtained across different machine-learning approaches. However, this similarity in performance did not fully translate into identical spatial predictions, as both models produced subtle but consistent and spatially structured differences, particularly in environmentally heterogeneous regions. This demonstrates that model choice has a measurable, although limited, influence on the spatial representation of soil properties and therefore matters for interpretation. An additional plausibility check indicated that both models capture expected land-use-related tendencies, although these effects remain weak at the considered scale. Overall, the study highlights that evaluating digital soil maps should extend beyond performance metrics by considering spatial prediction patterns, and it underscores the need for more transparent methodological reporting and code sharing to improve reproducibility and comparability across studies.

{{< pagebreak >}}

# Contributions

**Ulrike Krenz**: Conceptualization, Methodology, Investigation, Writing – original draft (Abstract, Introduction, Discussion, Conclusion), Writing – review & editing.

**Nina Preußler**: Conceptualization, Methodology, Investigation, Data curation and Documentation, Software, Formal analysis, Writing – original draft (Methods, Results), Writing – review & editing.

Both authors contributed equally to study design, interpretation of results, and manuscript preparation.

# Documentation of AI use

We used the tools listed below as supportive tools for the following purposes:

**ChatGPT 5.4**:  literature research support, support for structuring scientific argumentation, refine methodological design, coding assistant

**DeepL**: language support, text refinement

**Claude Sonnet 4.6**: coding assistant


{{< pagebreak >}}

# References

::: {#refs}
:::

{{< pagebreak >}}

# Appendix


```{r}
#| label: fig-plot-predictor-stack
#| fig-cap: "Spatial distribution of predictor variables."
#| fig-width: 12
#| fig-height: 16

# use only numeric variables
r_num <- predictor_stack[[1:14]]

# define units
units <- list(
  slope = "°",
  elevation = "m",
  aspect = "°",
  mean_temp = "°C",
  total_precip = "mm",
  latitude = "°",
  lc_dissim = "",
  clay = "g kg-1",
  silt = "g kg-1",
  ph = "",
  swc = "%",
  s_resp = "g C m-2 y-1",
  hfp = "",
  ndvi = ""
)

# define labels
labels <- list(
  slope = "slope",
  elevation = "elevation",
  aspect = "aspect",
  mean_temp = "mean temperature",
  total_precip = "total precipitation",
  latitude = "latitude",
  lc_dissim = "LC dissimilarity",
  clay = "clay content",
  silt = "silt content",
  ph = "pH",
  swc = "soil water content",
  s_resp = "soil respiration",
  hfp = "human footprint",
  ndvi = "NDVI"
)

plots <- vector("list", nlyr(r_num))

for (i in 1:nlyr(r_num)) {

  r <- r_num[[i]]
  nm <- names(r)

  unit <- units[[nm]]
  display_name <- labels[[nm]]

  title_text <- if (unit == "") display_name else paste0(display_name, "\n[", unit, "]")

  rmin <- minmax(r)[1]
  rmax <- minmax(r)[2]

  p <- ggplot() +
    geom_spatraster(data = r) +
    scale_fill_viridis_c(
      limits = c(rmin, rmax),
      breaks = c(rmin, rmax),
      na.value = "transparent",
      labels = round(c(rmin, rmax), 2)
    ) +
    labs(title = title_text, fill = NULL) +
    theme_void() +
    theme(
      plot.title = element_text(size = 10, hjust = 0.5),
      legend.position = "right"
    )

  plots[[i]] <- p
}

wrap_plots(plots, ncol = 4)
```

```{r}
#| label: tbl-accuracy-by-fold
#| tbl-cap: "Accuracy metrics by fold and by model"

accuracies <- accuracies %>%
  rownames_to_column("Fold") %>%
  mutate(Fold = as.integer(gsub("Fold", "", Fold))) %>%
  select(-Resample) %>%
  mutate(Model = ifelse(duplicated(Model), "", Model)) %>% # keep model name only in first row
  select(Model, everything()) # put Model as first column

knitr::kable(accuracies, digits = 3) 
```



