---
title: "Digital soil mapping of C:N ratio across Germany"
author: " & Nina Preußler"
format:
  html:
    self-contained: true 
execute: # set global chunk options
  echo: true # show code 
  warning: false # don't show warnings
  error: false # only render if no errors arise
bibliography: references.bib 
csl: https://www.zotero.org/styles/apa # citation style
editor: visual
knitr:
  opts_knit:
    root.dir: "~/Documents/00_ESDSRS/semester1/3_soil/soil-N-project" # working directory
---

# Title

Comparison of Random Forest and cubist machine learning for spatial prediction of soil C:N stoichiometry across Germany accuracy measures, spatial patterns, variable importance


# Abstract

max. 150 words

# Introduction



## Fragestellung

-   methodische Gap: keine Random Forest prediction in Deutschland

Schlagwörter: - Random Forest - C:N ratio Stöchiometrie - systematischer Vergleich - accuracy (Genauigkeit?) - spatial patterns (z.B. Boxplot nach land cover, elevation; visueller Eindruck, hotspots) - Variablen (importance, partial dependence)

How does spatial prediction of soil C:N stoichiometry across Germany differ when using a Random Forest model compared to a cubist approach?


# Methods

The analysis was conducted using R statistical software (REF) integrated in Quarto (REF). We extensively used the packages terra, caret, tidyverse/dplyr, and ggplot (REF).

```{r load-packages}
#| include: false

library(glcm)
library(Cubist)
library(raster) # required for dissimilarity
library(randomForest)
library(caret) # for k-fold cross-validation
library(sf)
library(gstat)
library(terra)
library(tmap)
library(devtools)
library(geodata)
library(spatstat) # for computing distance between sampling points
library(ggplot2)
library(patchwork)
library(tidyterra)
library(tidyverse)
library(glue)
options(renv.config.startup.quiet = TRUE)
```

The prediction of soil C:N was conducted on the scale of Germany. The shapefile of the administrative boundaries was obtained from `rnaturalearth` [@rnaturalearth].

```{r load-germany-mask}

library(rnaturalearth)
library(devtools)
devtools::install_github("ropensci/rnaturalearthhires", quiet = TRUE)

# get Germany boundary polygon
germany <- ne_countries(scale = 10, country = "Germany", returnclass = "sf") %>%
  select(geometry)

# reproject to UTM 32N
crs <- 'EPSG:25832'
germany_4326 <- germany
germany <- st_transform(germany, crs)

# # load land cover data to analyze distribution
# land_cover <- rast('data/covariates/land-cover/PROBAV_LC100_global_v3.0.1_2015-base_Discrete-Classification-map_EPSG-4326.tif') %>%
#   crop(germany_4326) %>%
#   project(crs, method='near')

```

Germany has a highly heterogenous landscape...
XXX % cropland
XXX % urban
temperate climate

## Data

### LUCAS soil survey

```{r load-lucas-data}

# load LUCAS topsoil data
lucas_eu <- st_read('data/LUCAS2015_topsoildata_20200323/LUCAS_Topsoil_2015_20200323-shapefile/LUCAS_Topsoil_2015_20200323.shp')

# reproject
lucas_eu <- st_transform(lucas_eu, crs)

# subset to Germany
lucas <- st_filter(lucas_eu, germany)
```

```{r calculate-sampling-point-distance}

# prepare point pattern with projection (required for distance)
lucas_ppp <- as.ppp(lucas)

# find distance to nearest neighbor for each point
distances <- nndist(lucas_ppp)

# summary statistics (in km2)
mean_dist <- mean(distances)/1000
sd_dist <- sd(distances)/1000
min_dist <- min(distances)/1000
max_dist <- max(distances)/1000
```

The measurements of soil properties were obtained from the LUCAS topsoil survey 2015 (CITE). In Germany, it consisted of `r nrow(lucas)` samples. Sampling points had a mean distance of `r round(mean_dist,2)` ± `r round(sd_dist,2)`, with minimum and maximum distances of `r round(min_dist,2)` and `r round(max_dist,2)`, respectively. This data was used to calculate the C:N ratios by dividing soil organic carbon (SOC) by total nitrogen (N).

```{r calculate-cn-ratio}
#| fig-cap: "Distribution of C:N ratio values from LUCAS 2015 samples in Germany."
#| fig-height: 3
#| fig-dpi: 200

# first make columns numeric, then add new column with ratio
lucas <- lucas %>%
  mutate(N = if_else(N == "< LOD", 0, as.numeric(N)),
         OC = if_else(OC == "< LOD", 0, as.numeric(OC)),
         cn_ratio = OC / N)

# summary statistics
cn_mean <- mean(lucas$cn_ratio)
cn_median <- median(lucas$cn_ratio)
cn_sd <- sd(lucas$cn_ratio)

ggplot(lucas, aes(cn_ratio)) +
  geom_histogram(binwidth = 0.5) +
  labs(x = "C:N ratio", y = "number of samples") +
  annotate(geom = "text", 
           x = 22, 
           y = 190, 
           label = paste('mean:', round(cn_mean,2)),
           hjust = 0, vjust = 1) + 
  annotate(geom = "text", 
           x = 22, 
           y = 165, 
           label = paste('median:', round(cn_median,2)),
           hjust = 0, vjust = 1) +
  annotate(geom = "text", 
           x = 22, 
           y = 140, 
           label = paste('standard deviation:', round(cn_sd,2)),
           hjust = 0, vjust = 1)

```

The distribution of C:N ratios obtained this way was strongly right-skewed, with most values below 10 (Fig. XXX). There was one notable outlier at a ratio of almost 40, but we did not remove it because it helps increases the prediction range of the random forest algorithm. The extreme value most likely did not arise from measuring error since a comparable value can also be seen in the later LUCAS 2018 survey.


## Covariates

The selection of environmental predictors was largely based on those proposed by @khosravani_soil_2026, who had identified them according to expert knowledge. We slightly modified them when not avoidable due to lacking documentation in the paper or data availability. An overview of the used predictors can be found in Table XXX and in the following we describe how we obtained and processed all predictors more in detail.


| Abbreviation | Full name | Unit | Source |
|----|----|----|----|
| ann_temp | Mean annual temperature |  |  |
| ann_precip |  |  |  |
| slope | Slope |  | https://zenodo.org/records/1447210 |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |

: Covariates used in C:N prediction, as proposed by Khosravani et al. (2025)

Elevation data at a resolution of 250 m was obtained from the global Digital Elevation Model provided by (REF Hengl, https://zenodo.org/records/1447210). Since it matched the target resolution, this raster was chosen as the reference to which we reprojected all other layers. terra [@terra] was used  to derive the covariates slope, aspect, and latitude from this elevation data.

The covariate layers temperature and precipitation were based on WorldClim [REF], which provides global climatic data at a 1km (??) resolution based on long-term monthly averages between 19XX and 20XX. We computed the mean annual temperature (ºC) and the total annual precipitation (mm).

```{r load-covariate-data-1}

# load DEM raster
dem <- rast('data/covariates/elevation/dtm_elevation_merit.dem_m_250m_s0..0cm_2017_v1.0.tif') %>%
  crop(germany_4326)

# calculate slope 
slope <- terrain(dem, v = "slope", unit = "degrees")
slope <-  project(slope, crs, method='bilinear') 

# calculate aspect
aspect <- terrain(dem, v = "aspect", unit = "degrees")

# latitude
latitude <- rast(list(latitude = init(dem, "y")))

# WorldClim (temperature and precipitation)
temp <- worldclim_country(country='germany', var='tavg', path='data/covariates/worldclim', res=0.5)
precip <- worldclim_country(country='germany', var='prec', path='data/covariates/worldclim', res=0.5)

temp_annual <- mean(temp) # mean annual temperature
precip_annual <- sum(precip) # total annual precipitation
```

Land cover dissimilarity was based on the land cover classification by https://zenodo.org/records/3939038 (REF), which contains the 2015 global land cover at a 100 m resolution. To reduce computational demand, it was reprojected to the target resolution of 250 m via mode resampling. With this raster, we computed the statistic "dissimilarity" from image texture measures from grey-level co-occurrence matrices (GLCM), calculating dissimilarity across all direction using a 9x9 window (capturing the dissimilarity in a 2.25 km$^2$ square).

```{r lc-and-dissimilarity}

# set option to redo the analysis (takes a few minutes) or load existing raster
redo_lc_dissim <- FALSE

# set path
lc_dissim_path <- "data/covariates/lc_dissim.tif"

# load land cover data and crop to Germany
land_cover <- rast('data/covariates/land-cover/PROBAV_LC100_global_v3.0.1_2015-base_Discrete-Classification-map_EPSG-4326.tif') %>%
  crop(germany_4326)


# obtain LC dissimilarity
# check if file exists AND redo flag is FALSE --> just load existing raster
if (file.exists(lc_dissim_path) && !redo_lc_dissim) {
  
  lc_dissim <- rast(lc_dissim_path)
  
} else {
  
  # reproject LC to target resolution (100 -> 250 m) to speed up processing
  land_cover_250m <- project(land_cover, slope, method='mode') # via majority voting

  # transform to RasterLayer (from raster package) - required for glcm
  land_cover_rast <- raster(land_cover_250m)
  
  # dissimilarity
  dissim_rast <- glcm(
    land_cover_rast,
    window = c(9,9), # take the 9x9 pixels forming the square around current pixel -> dissim in 1.125 km2 radius)
    shift = list(c(1,0), c(0,1), c(1,1), c(-1,1)), # average 4 directions (up, down, left, right)
    statistics = "dissimilarity"
    )
  
  # convert back to terra
  lc_dissim <- rast(dissim_rast)
  
  # save
  writeRaster(lc_dissim, lc_dissim_path, overwrite=TRUE)
}

```

Clay and silt content very poor documentation in paper! there it says from ISRIC; eventually downloaded from SoilGrids (250m resolution) - clay content (g kg-1) - in topsoil ( 0 to 5 cm)

For clay and silt, it was not specified which depth.

We therefore calculated weigthed averages from 0-5 and 5-15 cm values, the depths best matching the LUCAS samples which have a depth up to 20 cm.

Soil water content (SWC) Vol. water content at -33 kPa from SoilGrids (at field capacity) soilgrids

Soil respiration allegedly from ISRIC, but only found link there to https://www.earthdata.nasa.gov/data/catalog/ornl-cloud-soilresp-heterotrophicresp-1928-1 in supplementary materials in g CO2 per m2, but in NASA data g C per m2 per y


A global human footprint dataset with a resolution of 1 km was used from https://datadryad.org/dataset/doi:10.5061/dryad.3tx95x6d9.

2015 human footprint
https://doi.org/10.6084/m9.figshare.16571064) [MU et al]

The Normalized Difference Vegetation Index (NDVI) is an indicator of greenness. 
NDVI - not specified in paper how exactly calculated (besides GEE) - We used a 3-year growing season median (WHEN TO WHEN?) to capture long-term trends and avoid overrepresentation of weather influences.

```{r load-covariate-data-2}

# for next covariates, take weighted average of 0-5 cm and 5-15 cm raster
average_rasters <- function(rast1_path, rast2_path, weight1 = 1/3, weight2 = 2/3, resampling='bilinear') { 
  
  # load and crop first
  rast1 <- rast(rast1_path) %>%
    crop(germany_4326)
    
  # load and crop second, make sure to match grid of first via resampling
  rast2 <- rast(rast2_path) %>%
    crop(germany_4326) %>%
    project(rast1, method='bilinear')
  
  averaged_raster <- weight1 * rast1 + weight2 * rast2
  
  return(averaged_raster)
}

clay <- average_rasters('data/covariates/clay_0-5.tif', 'data/covariates/clay_5-15.tif')
silt <- average_rasters('data/covariates/silt_0-5.tif', 'data/covariates/silt_5-15.tif') 
ph <- average_rasters('data/covariates/ph_0-5.tif', 'data/covariates/ph_5-15.tif') / 10
# soil water content (SWC)
swc <- average_rasters('data/covariates/swc_0-5.tif', 'data/covariates/swc_5-15.tif')

# soil respiration
s_resp <- rast('data/SoilResp_HeterotrophicResp_1928/data/soil_Rs_mean.tif') %>%
  crop(germany_4326)

# human footprint (HFP)
hfp <- rast('data/covariates/hfp2015.tif') 
hfp <- crop(hfp, st_transform(germany, crs(hfp)))

# Normalized Difference Vegetation Index (NDVI)
ndvi <- rast('data/covariates/NDVI_Germany_2013_2015.tif')

```

Once we had obtained the raster data for all predictors, we projected them to the same CRS (EPSG 25832, UTM 32N) and resolution using bilinear resampling. The reference raster chosen for this alignment was the slope since it already had the target resolution of 250 m.

To account for differences in land cover, we encoded four natural land cover types as dummy variables: cropland, forest, grassland (herbaceous), and cropland.

From the resulting raster stack, we removed pixels corresponding to unsuitable land cover types (water bodies, snow, built-up land).

```{r prepare-raster-stack}

# same CRS and aligned

# set slope raster as template to align all other rasters with (because it has target resolution 250m)
slope <-  project(slope, crs, method='bilinear') 

# add to stack
predictor_stack <- c(slope)

# realign all remaining raster layers
predictors <- list(dem, aspect, temp_annual, precip_annual, latitude, lc_dissim, clay, silt, ph, swc, s_resp, hfp, ndvi)
for (predictor in predictors) {
  predictor <- project(predictor, slope, method="bilinear")
  predictor_stack <- append(predictor_stack, predictor)
}

# reproject land cover (categorical, therefore use different reprojection method)
land_cover_reproj <- project(land_cover, slope, method='mode') # majority voting
# predictor_stack <- append(predictor_stack, land_cover_reproj)
# make factor and add to stack
# land_cover_reproj_fact <- as.factor(land_cover_reproj)
# predictor_stack <- append(predictor_stack, land_cover_reproj_fact)


# # mask out certain land cover types
# unsuitable_lc_classes <- c(
#   0, # no input data available
#   50, # urban/built-up
#   70, # snow and ice
#   80, # permanent water bodies
#   200 # open sea
# )
# suitable_lc <- app(predictor_stack$lc_class, fun=function(x){ x[x %in% unsuitable_lc_classes] <- NA; return(x)} ) # set pixels corresponding to those LC types to NA
# predictor_stack <- mask(predictor_stack, suitable_lc) # set to NA in whole stack


# add dummy variables encoding major land cover types
forest_codes <- c(111,112,113,114,115,116,121,122,123,124,125,126)
forest <- land_cover_reproj %in% forest_codes  # returns binary raster (True/False)
grassland <- land_cover_reproj == 30
wetland <- land_cover_reproj == 90
cropland <- land_cover_reproj == 40

# add to predictor stack (multiply by one to make True to 1 and False to 0)
predictor_stack <- c(predictor_stack, forest*1, grassland*1, wetland*1, cropland*1)
# names(predictor_stack) <- c(names(predictor_stack)[1:(length(names(predictor_stack))-4)], 'forest', 'grassland', 'wetland', 'cropland')

# adjust names
names(predictor_stack) <- c('slope', 'elevation', 'aspect', 'mean_temp', 'total_precip', 'latitude', 'lc_dissim', 'clay', 'silt', 'ph', 'swc', 's_resp', 'hfp', 'ndvi', 'forest', 'grassland', 'wetland', 'cropland')

# mask with Germany (extract exact shape)
predictor_stack <- mask(predictor_stack, germany)

```

## Random Forest model

```{r prepare-model-data}

# extract data for each point from raster stack
predictor_values <- terra::extract(predictor_stack, lucas)[,-1] # remove ID

# combine with C:N data and remove unnecessary colums
data <- cbind(lucas['cn_ratio'], predictor_values) %>%
  drop_na() %>% 
  # select(!lc_class2) %>%
  st_drop_geometry() 

nrow(data)
```


We used k-fold cross-validation (CV) for hyperparameter tuning and estimating accuracy metrics. , to obtain more robust estimates. In k-fold CV, the dataset is split into k approximately equal parts ("folds"), and a model is iteratively fit on k-1 folds while the remaining fold is used for validation. The model accuracy metrics are then obtained by averaging the accuracy metrics for predictions (on the validation fold) across the folds. This CV approach offers a more robust estimate of model accuracy than a single training and testing split.

```{r rf-cross-validation}

set.seed(123)

# 10-fold CV (split model into 10 folds, run 10 times, each time, 1 fold is test while remaining 9 are used for training)
train_control <- trainControl(method = "cv", 
                              number = 10,
                              savePredictions = "final"
                              )


rfModel <- train(cn_ratio ~ ., # C:N ratio as a function of all predictors
                 data = data, 
                 method = "rf", # random forest
                 tuneLength = 5,
                 ntree = 500,
                 trControl = train_control, # implement CV as defined above
                 importance = TRUE) # compute permutation importance

rfModel
```

We implemented this k-fold CV using the `caret` \[@caret} package in R, setting k to 10. Caret automatically conducted hyperparameter tuning for the parameter `mtry`, i.e., the number of predictors used at each split, and found the lowest Root Mean Squared Error (RMSE) for `mtry` = `r rfModel$bestTune$mtry`. The number of trees was set to 500.

## Cubist model

The same 10-fold CV was applied for the Cubist model.

```{r cubist-cross-validation}

set.seed(123)

cubModel <- train(cn_ratio ~ ., # C:N ratio as a function of all predictors
                 data = data, 
                 method = "cubist", 
                 trControl = train_control, # implement CV as defined above for Random Forest
                 importance = TRUE) # compute permutation importance

cubModel
```

Hyperparameter tuning showed that `r cubModel$bestTune$committees` committees and `r cubModel$bestTune$neighbors` neighbors led to the lowest RMSE.

## Variable importance

To obtain the relative variable importance scores, a random forest and a cubist model with the optimal hyperparameters were fit to the whole dataset. 

## Model performance

Following @khosravani_soil_2026, we evaluated the model accuracy based on four metrics:

-   coefficient of determination (R$^2$), which indicates how much of the variance can be explained by the model
-   root mean squared error (RMSE), which measures XXX
-   concordance correlation coefficient (CCC), which quantifies the agreement between observed and predicted values
-   Bias, which shows whether the model systematically over- or underestimates the target variable

We followed the approach of @zhou_prediction_2021 and calculated accuracy metrics as the mean from the 10-fold cross-validation predictions to have more robust estimates of model accuracy.

```{r calculate-accuracy-measures}

# extract observations, predictions, and fold columns from best-fit model
rf_preds <- rfModel$pred %>%
  select(pred, obs, Resample)

# calculate accuracy metrics per fold
rf_preds_accuracy <- rf_preds %>%
  group_by(Resample) %>%
  summarise(
    RMSE = sqrt(mean((obs - pred)^2)),
    R2 = 1 - ((mean((obs - pred)^2))/(mean((obs - mean(rf_preds$obs))^2))),
    Bias = mean(obs - pred),
    CCC = (2 * cor(rf_preds$obs, rf_preds$pred) * sd(rf_preds$obs) * sd(rf_preds$pred)) /
  (var(rf_preds$obs) + var(rf_preds$pred) + (mean(obs) - mean(pred))^2)
)

# same for cubist
cubist_preds <- cubModel$pred %>%
  select(pred, obs, Resample)

cubist_preds_accuracy <- cubist_preds %>%
  group_by(Resample) %>%
  summarise(
    RMSE = sqrt(mean((obs - pred)^2)),
    R2 = 1 - ((mean((obs - pred)^2))/(mean((obs - mean(cubist_preds$obs))^2))),
    Bias = mean(obs - pred),
    CCC = (2 * cor(cubist_preds$obs, cubist_preds$pred) * sd(cubist_preds$obs) * sd(cubist_preds$pred)) /
  (var(cubist_preds$obs) + var(cubist_preds$pred) + (mean(obs) - mean(pred))^2)
)

# combine in one df
rf_preds_accuracy$Model <- "Random Forest"
cubist_preds_accuracy$Model <- "Cubist"
accuracies <- rbind(rf_preds_accuracy, cubist_preds_accuracy)

# average within each model across folds
accuracies_summarized <- accuracies %>%
  group_by(Model) %>%
  # summarise(across(everything(), mean)) %>%
  summarise(
    across(where(is.numeric), mean, .names = "{.col}_mean"),
    across(where(is.numeric), sd,   .names = "{.col}_sd")
  ) %>%
  # summarize in one column with ±
  transmute(
    Model = Model,
    RMSE = glue("{round(RMSE_mean, 2)} ± {round(RMSE_sd, 2)}"),
    R2 = glue("{round(R2_mean, 3)} ± {round(R2_sd, 2)}"),
    Bias = glue("{round(Bias_mean, 2)} ± {round(Bias_sd, 2)}"),
    CCC = glue("{round(CCC_mean, 2)} ± {round(CCC_sd, 3)}")
  )
  
```

To evaluate whether the models' performance was significantly different, we conducted a paired T-test.

```{r paired-t-test}
t_test_r2 <- t.test(accuracies$R2 ~ accuracies$Model)

t_test_rmse <- t.test(accuracies$RMSE ~ accuracies$Model)
```

## Mapping

To obtain mean and uncertainty maps, we used a bootstrapping approach as @khosravani_soil_2026 did. Due to computational constraints, we used 30 iterations for this process. In each iteration, a cubist and a random forest model were trained on a bootstrapped sample of the data (representing apprximately XXX% of samples) and then used to make spatial predictions based on the predictor stack. From the resulting 30 maps per model, we calculated the mean predicted C:N as well as uncertainty map based on the standard deviation. To investigate possible spatial patterns between the RF and Cubist predictions, we also computed a difference map by substracting the mean cubist from the mean RF map.

```{r bootstrapping}

# TAKES LONG (around 4 hours)! only redo if something about data changed - otherwise files will be loaded from disk (if available)
redo_bootstrap <- FALSE

# set number of bootstrapping runs
b_max = 30

n = nrow(data)

if (redo_bootstrap) {
  
  # downsample raster stack for testing
  # predictor_stack_coarse <- terra::aggregate(predictor_stack, fact = 3)

  for (b in seq(1,b_max)) {
    
    # progress report
    cat("Bootstrap iteration:", b, "of", b_max, "\n")
    
    # get indexes via bootstrap sampling (with replacement)
    idxs <- sample(n, size=n, replace = TRUE)
    
    # subset data
    boot_data <- data[idxs,]
    
    # train models (with best hyperparameter tuning values) on it
    rf_model <- randomForest(
      cn_ratio ~ .,      # C:N as function of all other columns (predictors)
      data = boot_data,
      mtry = rfModel$bestTune$mtry,
      ntree = 500       # number of trees
      )
    cubist_model <- cubist(
      x = select(boot_data, !cn_ratio), 
      y = boot_data$cn_ratio,
      committees = cubModel$bestTune$committees
      )
    
    # spatial prediction
    map_rf <- terra::predict(
      object = predictor_stack, # what to predict from
      model  = rf_model,
      cores=4, # parallel processing
      cpkgs = "randomForest", # package needed to run prediction
      filename = paste0('data/bootstrap_prediction_layers/rf_', b, '.tif' ), # write to disk instead of keeping in memory
      overwrite = TRUE
      )
    map_cubist <- terra::predict(
      object = predictor_stack, 
      model  = cubist_model,
      cores=4,
      cpkgs = "Cubist", 
      filename = paste0('data/bootstrap_prediction_layers/cub_', b, '.tif' ), 
      overwrite = TRUE
    )
  }
}

```

```{r}

# load bootstrapping map predictions
path <- 'data/bootstrap_prediction_layers/'
rf_boot_stack <- rast(paste0(path, list.files(path, pattern='rf*')))
cub_boot_stack <- rast(paste0(path, list.files(path, pattern='cub*')))

# mask cubist stack with Germany (because it also predicts for NA values)
cub_boot_stack <- mask(cub_boot_stack, germany)

# compute mean and standard deviation across layers
rf_mean_map <- app(rf_boot_stack, mean)
rf_sd_map   <- app(rf_boot_stack, sd)

cub_mean_map <- app(cub_boot_stack, mean)
cub_sd_map   <- app(cub_boot_stack, sd)

# difference map
diff_map <- rf_mean_map - cub_mean_map
diff_map_abs <- abs(diff_map)

```

e.g., https://www.sciencedirect.com/science/article/pii/S0048969720361908 created 100 maps

## Ecological plausibility

To illustrate a possible use case for C:N maps, we tested whether predicted C:N ratios in grasslands were significantly higher if they were located in protected areas. 

For this purpose, we used a shapefile of Natura 2000 areas obtained from https://www.eea.europa.eu/en/datahub/datahubitem-view/6fc8ad2d-195d-40f4-bdec-576e7d1268e4.

We used a separate grassland map from XXX since the land cover we used did not have it as a distinct class.

```{r prepare-natura-grassland-df}

# load Natura 2000 shapefile
natura_vect <- vect('data/SHP/Natura2000_end2024_epsg3035.shp') %>%
  project(crs) %>%
  crop(germany)

# rasterize
natura <- rasterize(natura_vect, slope, background=0, field=1)

# load grassland raster
grassland_eu <- rast('data/GRA_2015_100m_eu_03035_V1_2.tif')

# crop
grassland_ger <- grassland_eu %>%
  crop(st_transform(germany, crs(grassland_eu)))

# make binary raster (1 where grassland, 0 all others)
grassland_ger <- grassland_ger == 1

# reproject to match raster stack grid (250 m) using mode resampling --> only classify as grassland if majority of 100 pixels are grassland 
grassland_ger <- project(grassland_ger, slope, method='mode')

# function to mask rasters by grassland and natura, and convert to dataframe
# function developed by Claude Sonnet 4.6
create_cn_boxplot_df <- function(cn_raster1, cn_raster2,
                                  grassland_raster, natura_raster,
                                  model_names = c("model1", "model2")) {
  
  # Stack all rasters and extract values in one pass
  stack <- c(cn_raster1, cn_raster2, grassland_raster, natura_raster)
  vals  <- terra::values(stack, dataframe = TRUE, na.rm = FALSE)
  colnames(vals) <- c("cn1", "cn2", "grassland", "natura")
  
  # Build a valid-cell mask (exclude cells where any layer is NA)
  valid <- complete.cases(vals)
  vals  <- vals[valid, ]
  
  # Pivot the two CN columns into long format
  df <- bind_rows(
    data.frame(
      grassland = as.logical(vals$grassland),
      protected = as.logical(vals$natura),
      model     = model_names[1],
      cn_ratio  = vals$cn1
    ),
    data.frame(
      grassland = as.logical(vals$grassland),
      protected = as.logical(vals$natura),
      model     = model_names[2],
      cn_ratio  = vals$cn2
    )
  )
  
  df$model <- as.character(df$model)
  return(df)
}

grass_natura_df <- create_cn_boxplot_df(
  cn_raster1       = rf_mean_map,
  cn_raster2       = cub_mean_map,
  grassland_raster = grassland_ger,
  natura_raster    = natura,
  model_names      = c("RF", "Cubist")
) %>%
  filter(grassland == TRUE)
```

# Results

## Model performance

-   models had very similar performance
-   RF slightly higher R2 and lower RMSE, see @accuracy-table
-   RF tends to overestimate values (slightly negative bias), especially lower observed values
-   cubist tends to underestimate (positive bias)
-   cubist had higher CCC, also visible in @plot-obs-vs-pred
-   

The paired T-tests revealed that the two models did not differ significantly in their R$^2$ (p = XXX) or RMSE (p = XXX).

```{r accuracy-table}
#| tbl-cap: "Accuracy metrics of the two models. Values indicate mean and standard deviation from the 10-fold cross-validation."

knitr::kable(accuracies_summarized)
```

```{r plot-obs-vs-pred}
#| fig-cap: "Observed vs. predicted values for Random Forest and Cubist models. Predicted values are from out-of-box prediction during the 10-fold CV"

# two-panel plot with RF and cubist

# Bin size control + color palette
p1 <- ggplot(rf_preds, aes(obs, pred) ) +
  geom_hex(bins = 70) +
  scale_fill_continuous(limits=c(0,30), type = "viridis", name = 'count') +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  coord_equal() 

p2 <- ggplot(cubist_preds, aes(obs, pred) ) +
  geom_hex(bins = 70) +
  scale_fill_continuous(limits=c(0,30), type = "viridis", name = 'count') +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  coord_equal() #+
  # xlim(5, 20) +
  # ylim(5,20)

(p1 / p2) + plot_layout(guides = "collect")

```

```{r correlation-rf-cubist}
corr_rf_cub <- cor(arrange(rfModel$pred, rowIndex)['pred'], arrange(cubModel$pred, rowIndex)['pred'])
```

Random forest and cubist predictions had a correlation coefficient of `r round(corr_rf_cub,3)`.

## Maps

```{r mean-maps}

# Determine shared limits
zlim <- range(c(values(rf_mean_map), values(cub_mean_map)), na.rm = TRUE)

p1 <- ggplot() +
  geom_spatraster(data = rf_mean_map) +
  scale_fill_viridis_c(limits = zlim, name = "C:N") +
  ggtitle("Random Forest") +
  theme_minimal()

p2 <- ggplot() +
  geom_spatraster(data = cub_mean_map) +
  scale_fill_viridis_c(limits = zlim, name = "C:N") +
  ggtitle("Cubist") +
  theme_minimal()

# Combine with patchwork, collecting the shared legend
(p1 | p2) + plot_layout(guides = "collect")
```

```{r uncertainty-maps}

# Determine shared limits
zlim <- range(c(values(rf_sd_map), values(cub_sd_map)), na.rm = TRUE)

p1 <- ggplot() +
  geom_spatraster(data = rf_sd_map) +
  scale_fill_viridis_c(limits = zlim, name = "C:N standard deviation") +
  ggtitle("Random Forest") +
  theme_minimal()

p2 <- ggplot() +
  geom_spatraster(data = cub_sd_map) +
  scale_fill_viridis_c(limits = zlim, name = "C:N standard deviation") +
  ggtitle("Cubist") +
  theme_minimal()

# Combine with patchwork, collecting the shared legend
(p1 | p2) + plot_layout(guides = "collect")

```

```{r}
# sd_rf_masked <- mask(rf_sd_map, suitable_lc)
mean_sd_rf <- mean(values(rf_sd_map), na.rm=TRUE)
sd_sd_rf <- sd(values(rf_sd_map), na.rm = TRUE)

# sd_cub_masked <- mask(cub_sd_map, suitable_lc)
mean_sd_cub <- mean(values(cub_sd_map), na.rm=TRUE)
sd_sd_cub <- sd(values(cub_sd_map), na.rm = TRUE)
```


With a mean standard deviation of `r round(mean_sd_rf, 2)` ± `r round(sd_sd_rf,2)`, random forest predictions were on average less uncertain than those from cubist, which had a mean standard deviation of `r round(mean_sd_cub, 2)` ± `r round(sd_sd_cub,2)`.

Most extreme standard deviation in the Cubist map were located where the random forest did not make predictions due to missing data, mostly where large water bodies are found.


```{r plot-difference-map}
#| fig-cap: "Difference map"

# Define breakpoints and colors
breaks <- c(minmax(diff_map)[1], -1, 1, minmax(diff_map)[2])

colors <- c("red",        # below -2
            "lightgrey",  # between -2 and 2
            "blue")       # above 2

# Plot with custom colors
plot(diff_map, breaks = breaks, col = colors, legend = TRUE)

# ggplot() +
#   geom_spatraster(data = diff_map) +
#   scale_fill_viridis_c(name = "C:N difference") +
#   ggtitle("Difference map") +
#   theme_minimal()

```
 
 Red (negative values) indicates that cubist prediction > RF prediction, whereas blue (positive values) indicates higher RF predictions.

A simple correlation matrix (see Appendix) showed no large correlation between either relative(??? with + or -) or absolute difference and any of the predictors. Absolute difference were moderately linearly correlated with standard deviation (r = 0.37 for RF and r = 0.27 for cubist).

```{r predictions-by-land-cover}

grassland_preds <- mask(rf_mean_map, subst(grassland, 0, NA))
forest_preds <- mask(rf_mean_map, subst(forest, 0, NA))
crop_preds <- mask(rf_mean_map, subst(cropland, 0, NA))
wetland_preds <- mask(rf_mean_map, subst(wetland, 0, NA))

boxplot(cbind(values(grassland_preds), values(forest_preds), values(crop_preds), values(wetland_preds)))

grassland_uncert <- mask(rf_sd_map, subst(grassland, 0, NA))
forest_uncert <- mask(rf_sd_map, subst(forest, 0, NA))
crop_uncert <- mask(rf_sd_map, subst(cropland, 0, NA))
wetland_uncert <- mask(rf_sd_map, subst(wetland, 0, NA))

```

grassland mean uncertainty `r mean(grassland_uncert, na.rm = TRUE)` ± `r sd(grassland_uncert, na.rm = TRUE)`

forest mean uncertainty `r mean(forest_uncert, na.rm = TRUE)` ± `r sd(forest_uncert, na.rm = TRUE)`

crop mean uncertainty `r mean(crop_uncert, na.rm = TRUE)` ± `r sd(crop_uncert, na.rm = TRUE)`

wetland mean uncertainty `r mean(wetland_uncert, na.rm = TRUE)` ± `r sd(wetland_uncert, na.rm = TRUE)`

```{r}

lucas_pred <- terra::extract(c(rf_mean_map, cub_mean_map, rf_sd_map, cub_sd_map), lucas)[,-1] %>%
  cbind(lucas[c('cn_ratio', 'LC0_Desc')]) %>%
  rename(
    rf_mean = mean,
    cub_mean = mean.1,
    rf_sd = sd,
    cub_sd = sd.1,
    land_cover = LC0_Desc
  ) %>%
  mutate(
    rf_resid = cn_ratio - rf_mean,
    cub_resid = cn_ratio - cub_mean
  )

lucas_pred_long <- lucas_pred %>%
  pivot_longer(
    cols         = c(rf_mean, cub_mean, rf_resid, cub_resid, rf_sd, cub_sd),
    names_to     = c("model", ".value"),
    names_pattern = "(.+)_(mean|resid|sd)"
  )

ggplot(lucas_pred_long, aes(x=model, y=mean, fill = model)) +
  geom_boxplot() +
  facet_wrap(~land_cover)

ggplot(lucas_pred_long, aes(x=model, y=sd, fill = model)) +
  geom_boxplot() +
  facet_wrap(~land_cover)

ggplot(lucas_pred_long, aes(x=model, y=resid, fill = model)) +
  geom_boxplot() +
  facet_wrap(~land_cover)

```

## Variable importance


```{r variable-importance}

# extract importance scores
var_imp_rf <- varImp(rfModel)$importance
var_imp_cub <- varImp(cubModel)$importance

# start new df with RF
imp_df <- tibble(
  variable = rownames(var_imp_rf),
  RF = var_imp_rf$Overall
) 
  
# add Cubist scores
imp_df <- tibble(
  variable = rownames(var_imp_cub),
  Cubist = var_imp_cub$Overall
  ) %>%
  full_join(imp_df, by = join_by(variable)) %>% # combine with RF scores
  pivot_longer( # modify format
    cols = c(RF, Cubist),
    names_to = 'model',
    values_to = 'importance'
  ) %>%
  arrange(desc(model), desc(importance))

# make variable into ordered factor column based on RF importance scores (to order plot by descending importance scores)
rf_order <- imp_df %>%
  filter(model == "RF") %>%
  arrange(importance) %>%
  pull(variable)
imp_df$variable <- factor(imp_df$variable, levels = rf_order)

ggplot(imp_df, aes(variable, importance, fill=model)) +
  geom_col(position='dodge') +
  coord_flip() +
  xlab(element_blank()) +
  ylab('Relative importance [%]')
```

Ph was by far the most important predictor in both models. NDVI was ranked second for both.

The least important numeric predictors were soil respiration and aspect. 



## Ecological plausibility

```{r compare-natura-boxplot}


ggplot(grass_natura_df, aes(protected, cn_ratio, fill = model)) +
  geom_boxplot() 

grass_natura_df_rf <- filter(grass_natura_df, grass_natura_df$model == 'RF')

# anova_rf <- aov(cn_ratio ~ protected, data = grass_natura_df_rf)
# summary(anova_rf)

# Mann–Whitney U
wilcox.test(cn_ratio ~ protected, data = grass_natura_df_rf)

# get effect size
library(effsize)
cliff.delta(cn_ratio ~ protected, data = df)

```



# Discussion

// Diskussion (methodisch + inhaltlich + Literaturvergleich + Limitationen)

## Model accuracy

-   slightly lower than reference paper for both models, even though we used the same input data. However, due to poor documentation, we Furthermore, the authors did not specify how they obtained their accuracy metrics, so they might only represent measures for only favorable training-testing split, while our metrics provide a more robust estimate due to th 10-fold cross-validation.

The ideal way of validating our map results would be through additional random or strategic sampling (Wadoux ref). This is often not possible due to cost and time constraints. For not strongly clustered points, cross-validation can also provide nearly unbiased estimates of model accuracy, though they never allow you to make certain inferences about the true population (you only stay in your sampled data) (wadoux ref).

## Environmental predictors

Identical to the results by @khosravani_soil_2026, pH was also by far the most important predictor in both our models.

A notable difference was the importance of NDVI. While @khosravani_soil_2026 had a relative importance of \<5% for this predictor, in our models, it was the second most important covariate. However, it needs to be noted that it was not clear how @khosravani_soil_2026 had computed the NDVI map. We wanted to capture it as a long-term indicator of productivity by using a 3-year growing-season median, which seemed to work well.

A controversial (?) and surprising choice by @khosravani_soil_2026 was to include land cover dissimilarity rather than land cover classes, which they did not explain nor document properly how it was derived. 
// What might LC dissimilarity really tell us? How could it be related to C:N ratio

*But does NDVI affect the C:N ratio directly or is it more likely to be a proxy for some other property?*

## Limitations

-   WorldClim data from 1970 to 2000 - climate change might already have had an effect by 2018
-   

# Conclusion

// Fazit (kurz, prägnant, 3–6 Sätze)

-   importance of accurate documentation and ideally sharing code for reproducibility

# AI documentation

## Preparation phase

-   get an overview of possible research directions
-   refine search terms

## Coding

- 

# Appendix

## Predictor maps

```{r}
plot(predictor_stack)
```


```{r plot-predictor-stack}
#| eval: false


# set up plot grid
par(mfrow = c(3, 6)) 

# define units
units <- list(
  slope = "°",
  elevation = "m",
  aspect = "°",
  mean_temp = "°C",
  total_precip = "mm",
  latitude = "°",
  lc_dissim = "none",
  clay = "g/kg",
  silt = "g/kg",
  ph = "none",
  swc = "???",
  r_resp = "???",
  hfp = "none",
  ndvi = "none"
)

for (i in 1:nlyr(predictor_stack)) {
  
  r <- predictor_stack[[i]]
  nm <- names(r)

  vals <- values(r)
  
  if (all(vals %in% c(0,1), na.rm = TRUE)) {
    # boolean layer
    plot(
      r,
      main = paste0(nm),
      col = c("lightgrey", "darkgreen"),
      breaks = c(-0.5, 0.5, 1.5)#,
      # plg = list(at = c(0,1), labels = c("False","True"))
    )
    
  } else {
    # continuous layer
    rmin <- minmax(r)[1]
    rmax <- minmax(r)[2]
    
    plot(
      r,
      main = paste0(nm, " [", units[nm], "]"),
      col = hcl.colors(50, "viridis"),
      plg = list(at = c(rmin, rmax), labels = round(c(rmin, rmax), 2))
    )
  }
}
```


## Spatial autocorrelation C:N

```{r c-n-experimental-variogram}

# use projection (in m)
lucas_proj <- st_transform(lucas, 'EPSG:25832')

# make gstat object
gstat_cn <- gstat(formula = cn_ratio ~ 1, data = lucas_proj)

vg_cn <- variogram(gstat_cn, cloud=F, boundaries=seq(3000, 30000, 3000))
plot(vg_cn, plot.nu = TRUE)

vg_cn <- variogram(gstat_cn, cloud=F, boundaries=seq(2000, 10000, 2000))
plot(vg_cn, plot.nu = TRUE)

```

--\> spatial autocorrelation only to distance of \~7.5km and only small fraction of points have other points within this distance (especially when trying variogram with smaller cloud classes), so geostatistical approaches such as kriging not ideal

## Correlation matrix

```{r explore-correlation}

corr_rasters <- c(predictor_stack, rf_mean_map, rf_sd_map, cub_mean_map, cub_sd_map, diff_map, diff_map_abs)

names(corr_rasters) <- c(names(predictor_stack), 'rm_mean', 'rf_sd', 'cub_mean', 'cub_sd', 'diff', 'abs_diff')

# correlation matrix
# cor_mat <- layerCor(corr_rasters, fun = "cor", na.rm = TRUE)

# sample to reduce processing time
corr_rasters_samp <- spatSample(corr_rasters, size = 1e4, method = "random", na.rm = TRUE, as.df = TRUE)

cor(corr_rasters_samp)

```

# References
