---
title: "Digital soil mapping of C:N ratio across Germany"
author: "Nina Preußler"
format:
  html:
    self-contained: true
editor: visual
knitr:
  opts_knit:
    root.dir: "~/Documents/00_ESDSRS/semester1/3_soil/soil-N-project"
---

# Title

Comparison of Random Forest and cubist machine learning for spatial prediction of soil C:N stoichiometry across Germany
accuracy measures, spatial patterns, variable importance

# Introduction

## Why is C:N important?

-   What does it tell you about the soil?

## How is C:N changing?

-   What processes and factors affect it?
-   EU/Germany trends

## Why DSM?

-   existing maps for EU/Germany
    -   method
    -   data
    -   quality
    -   challenges
    
## Fragestellung

- methodische Gap: keine Random Forest prediction in Deutschland

Schlagwörter:
- Random Forest
- C:N ratio Stöchiometrie 
- systematischer Vergleich
  - accuracy (Genauigkeit?)
  - spatial patterns (z.B. Boxplot nach land cover, elevation; visueller Eindruck, hotspots)
  - Variablen (importance, partial dependence)

How does spatial prediction of soil C:N stoichiometry across Germany differ when using a Random Forest model compared to a cubist approach? 

"ökologischer Plausibilitätscheck"


# Methods

```{r load-packages, include=FALSE}

library(glcm)
library(raster) # required for dissimilarity
library(randomForest)
library(caret) # for k-fold cross-validation
library(sf)
library(gstat)
library(terra)
library(tmap)
library(stars)
library(devtools)
library(geodata)
library(spatstat) # for computing distance between sampling points
library(ggplot2)
library(tidyverse)


```

## Study area

```{r load-germany-mask}

library(rnaturalearth)
library(devtools)
devtools::install_github("ropensci/rnaturalearthhires", quiet = TRUE)

# get Germany boundary polygon
germany <- ne_countries(scale = 10, country = "Germany", returnclass = "sf") %>%
  select(geometry)

# get bounding box
st_bbox(germany)

# reproject to UTM 32N
crs <- 'EPSG:25832'
germany_4326 <- germany
germany <- st_transform(germany, crs)
```

## Data

## LUCAS soil survey

LUCAS topsoil survey 2018

```{r load-lucas-data}

# load LUCAS topsoil data
lucas_eu <- st_read('data/LUCAS2015_topsoildata_20200323/LUCAS_Topsoil_2015_20200323-shapefile/LUCAS_Topsoil_2015_20200323.shp')

# reproject
lucas_eu <- st_transform(lucas_eu, crs)

# subset to Germany
lucas <- st_filter(lucas_eu, germany)

lucas %>% glimpse()
plot(lucas["N"])
```

### C:N ratio

calculated with $\frac{\text{total N}}{\text{organic C}}$

```{r calculate-cn-ratio}

# first make columns numeric, then add new column with ratio
lucas <- lucas %>%
  mutate(N = if_else(N == "< LOD", 0, as.numeric(N)),
         OC = if_else(OC == "< LOD", 0, as.numeric(OC)),
         cn_ratio = OC / N)

ggplot(lucas, aes(cn_ratio)) +
  geom_histogram(binwidth = 0.5) 
```

```{r remove-nans-and-outliers}

# copy before cleaning
lucas_init <- lucas

# summary statistics
cn_mean <- mean(lucas$cn_ratio)
cn_median <- median(lucas$cn_ratio)
cn_sd <- sd(lucas$cn_ratio)

iqr <- IQR(lucas$cn_ratio)
first_quantile <- quantile(lucas$cn_ratio)['25%']
third_quantile <- quantile(lucas$cn_ratio)['75%']

# remove nan and outliers (> IQR +- 1.5 * IQR)
lucas <- lucas %>%
  drop_na(cn_ratio) %>%
  mutate(cn_ratio_no_outlier = between(cn_ratio, first_quantile-1.5*iqr, third_quantile+1.5*iqr))

ggplot(lucas, aes(cn_ratio, fill=cn_ratio_no_outlier)) +
  geom_histogram(binwidth = 0.5) 
```
but doesn't really make sense because then we limit the prediction range of RF

```{r plot-lucas-points-germany}

tmap_mode("view")
tm_basemap('Esri.WorldImagery') +
  tm_shape(germany) +
  tm_lines(col="black", lwd = 3) +
  tm_shape(lucas) +
  tm_dots(fill="cn_ratio",
          size = 0.5,
          fill.scale = tm_scale_continuous(values = "cividis",
                                           limits = c(0, 20)))
```

#### Distance between sampling points

```{r}

# prepare point pattern with projection (required for distance)
lucas_ppp <- as.ppp(lucas)

# find distance to nearest neighbor for each point
distances <- nndist(lucas_ppp)

# summary statistics
print(paste('Total number of points:', length(lucas$cn_ratio)))
print(paste('Mean distance:', round(mean(distances)/1000,2), 'km'))
print(paste('Median distance:', round(median(distances)/1000,2), 'km'))
print(paste('Standard deviation distance:', round(sd(distances)/1000,2), 'km'))
print(paste('Minimum distance:', round(min(distances)/1000,2), 'km'))
print(paste('Maximum distance:', round(max(distances)/1000,2), 'km'))

```

## Covariates

based on Khosravani et al. (2025)

| Abbreviation | Full name | Unit | Source |
|----|----|----|----|
| ann_temp | Mean annual temperature |  |  |
| ann_precip |  |  |  |
| slope | Slope |  | https://zenodo.org/records/1447210 |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |

: Covariates used in C:N prediction, as proposed by Khosravani et al. (2025)

### Slope and aspect

Used DEM from Hengl / Yamazaki et al (https://zenodo.org/records/1447210).

```{r load-dem}

# load raster
dem_global <- rast('data/covariates/elevation/dtm_elevation_merit.dem_m_250m_s0..0cm_2017_v1.0.tif') 

# crop to Germany
dem <- crop(dem_global, germany_4326)

```

```{r calculate-slope-and-aspect}

# calculate slope 
slope <- terrain(dem, v = "slope", unit = "degrees")

# calculate aspect
aspect <- terrain(dem, v = "aspect", unit = "degrees")

```

calculated slope and aspect in degrees using terra

### Temperature and precipitation

-   loaded from WorldClim (XXX)

```{r load-worldclim-data}

# WorldClim (temperature and precipitation)
temp <- worldclim_country(country='germany', var='tavg', path='data/covariates/worldclim', res=0.5)
precip <- worldclim_country(country='germany', var='prec', path='data/covariates/worldclim', res=0.5)

temp_annual <- mean(temp) # mean annual temperature
precip_annual <- sum(precip) # total annual precipitation

```


### Latitude

extracted latitude from rasters

```{r extract-latitude}
latitude <- rast(list(latitude = init(dem, "y")))
```
### Land cover dissimilarity

link in paper was not accessible

instead https://zenodo.org/records/3939038

but is dissimilarity really more "aussagekraeftig" than the categorical LC type?
maybe try if adding LC type would increase accuracy?
```{r lc-dissimilarity}

# load land cover data and crop to Germany
land_cover_global <- rast('data/covariates/land-cover/PROBAV_LC100_global_v3.0.1_2015-base_Discrete-Classification-map_EPSG-4326.tif')
land_cover <- crop(land_cover_global, germany_4326)

plot(land_cover, col=map.pal('viridis',200))

# transform to RasterLayer (from raster package) - required for glcm
land_cover_rast <- raster(land_cover)

# dissimilarity (from ChatGPT)
dissim_rast <- glcm(
  land_cover_rast,
  window = c(3,3), # take only the 8 pixels forming the square directly around current pixel
  shift = list(c(1,0), c(0,1), c(1,1), c(-1,1)), # average 4 directions (up, down, left, right)
  statistics = "dissimilarity"
)

# convert back to terra
lc_dissim <- rast(dissim_rast)

plot(lc_dissim)

```


### Clay and silt content

very poor documentation in paper!
there it says from ISRIC

eventually downloaded from SoilGrids (250m resolution)
- clay content (g kg-1)

```{r load-silt-and-clay}

# load and crop to Germany

clay <- rast('data/covariates/clay_0-5.tif') %>%
  crop(germany_4326)
silt <- rast('data/covariates/silt_0-5.tif') %>%
  crop(germany_4326)


```

### pH

```{r load-ph}

ph <- rast('data/covariates/ph_0-5.tif')

```

### Soil water content (SWC)

Vol. water content at -33 kPa from SoilGrids (at field capacity)

```{r load-swc}

swc <- rast('data/covariates/swc.tif')
```

### Soil respiration

allegedly from ISRIC, but only found link there to https://www.earthdata.nasa.gov/data/catalog/ornl-cloud-soilresp-heterotrophicresp-1928-1

in supplementary materials in g CO2 per m2, but in NASA data g C per m2 per y

```{r}
s_resp <- rast('data/SoilResp_HeterotrophicResp_1928/data/soil_Rs_mean.tif') %>%
  crop(germany_4326)
```


### Extra: Microbial biomass

```{r load-microbial-biomass-data}

microb <- rast('data/GEB_maps/Cmic_gen_annual_mean_t.tif')
microb_eu <- read_stars('data/GEB_maps/Cmic_gen_annual_mean_t.tif')

# subset to
#microb <- st_filter(microb_eu, germany)

plot(microb_eu)
```


## Random Forest model

```{r prepare-raster-stack}

# same CRS and aligned

# set slope raster as template to align all other rasters with (because it has target resolution 250m)
slope <-  project(slope, crs, method='bilinear') 

# add to stack
predictor_stack <- c(slope)

# realign all remaining raster layers
predictors <- list(aspect, temp_annual, precip_annual, latitude, lc_dissim, clay, silt, ph, swc, s_resp)
for (predictor in predictors) {
  predictor <- project(predictor, slope, method="bilinear")
  predictor_stack <- append(predictor_stack, predictor)
}

# adjust names
names(predictor_stack) <- c('slope', 'aspect', 'mean_temp', 'total_precip', 'latitude', 'lc_dissim', 'clay', 'silt', 'ph', 'swc', 's_resp')

# mask with Germany (extract exact shape)
predictor_stack <- mask(predictor_stack, germany)

plot(predictor_stack)

```

```{r split-train-test}

set.seed(123)

# extract data for each point from raster stack
predictor_values <- terra::extract(predictor_stack, lucas)[,-1] # remove ID

# add extracted values to C:N ratio data
data <- cbind(lucas[c('Point_ID', 'cn_ratio')], predictor_values) %>%
  st_drop_geometry()

# randomly select points by proportion
train_data <- data %>%
  slice_sample(prop = 0.7) 
  
train_ids <- train_data$Point_ID

# use remaining rows for test
test_data <- data %>%
  filter_out(Point_ID %in% train_ids) %>% # keep only IDs NOT in training data
  select(!Point_ID)

```

In the paper, they used a bootstrapping approach, but did not specify with how many repeats.
```{r prepare-cross-validation}
train_control <- trainControl(method = "cv", number = 10)

```

```{r train-random-forest-model}

set.seed(123)

# remove Point_ID
train_data <- select(train_data, !Point_ID)

# train model
rf_model <- randomForest(
  cn_ratio ~ .,      # C:N as function of all other columns (predictors)
  data = train_data,
  ntree = 500,       # number of trees
  importance = TRUE  # calculate variable importance scores
)

```

```{r test-model}

train_pred <- predict(
  rf_model,
  newdata = train_data,
)

plot(train_data$cn_ratio, train_pred)
abline(0, 1, col='red')
```

```{r spatial-prediction}

cn_map <- terra::predict(
  object = predictor_stack, # what to predict from
  model  = rf_model
)

names(cn_map) <- c("Mean predicted C:N ratio")

plot(cn_map)

```


## Model evaluation

-   coefficient of determination (R2)
-   concordance correlation coefficient (CCC)
-   root mean squared error (RMSE)
-   Bias

```{r calculate-accuracy-measures}


```

### Uncertainty map

## Ecological plausibility

- Natura 2000

# Results

## Exploratory data analysis

### C:N ratio

#### Spatial autocorrelation

```{r c-n-experimental-variogram}

# use projection (in m)
lucas_proj <- st_transform(lucas, 'EPSG:25832')

# make gstat object
gstat_cn <- gstat(formula = cn_ratio ~ 1, data = lucas_proj)

vg_cn <- variogram(gstat_cn, cloud=F, boundaries=seq(3000, 30000, 3000))
plot(vg_cn, plot.nu = TRUE)

vg_cn <- variogram(gstat_cn, cloud=F, boundaries=seq(2000, 10000, 2000))
plot(vg_cn, plot.nu = TRUE)

```
--> spatial autocorrelation only to distance of ~7.5km and only small fraction of points have other points within this distance (especially when trying variogram with smaller cloud classes), so geostatistical approaches such as kriging not ideal

## Variable importance

```{r plot-variable-importance}
varImpPlot(rf_model, main = "Predictor Importance")

```

## Evalutation

### Summary statistics of predicted C:N by land cover class

# Discussion

## Limitations

-   WorldClim data from 1970 to 2000 - climate change might already have had an effect by 2018

### Microbial

------------------------------------------------------------------------

# AI documentation

## Preparation phase

-   get an overview of possible research directions
-   refine search terms
