---
title: "Digital soil mapping of C:N ratio across Germany"
author: "Nina Preußler"
format:
  html:
    self-contained: true
bibliography: references.bib
csl: apa
editor: visual
knitr:
  opts_knit:
    root.dir: "~/Documents/00_ESDSRS/semester1/3_soil/soil-N-project"
---

# Title

Comparison of Random Forest and cubist machine learning for spatial prediction of soil C:N stoichiometry across Germany
accuracy measures, spatial patterns, variable importance

# Introduction

## Why is C:N important?

-   What does it tell you about the soil?

## How is C:N changing?

-   What processes and factors affect it?
-   EU/Germany trends

## Why DSM?

-   existing maps for EU/Germany
    -   method
    -   data
    -   quality
    -   challenges
    
## Fragestellung

- methodische Gap: keine Random Forest prediction in Deutschland

Schlagwörter:
- Random Forest
- C:N ratio Stöchiometrie 
- systematischer Vergleich
  - accuracy (Genauigkeit?)
  - spatial patterns (z.B. Boxplot nach land cover, elevation; visueller Eindruck, hotspots)
  - Variablen (importance, partial dependence)

How does spatial prediction of soil C:N stoichiometry across Germany differ when using a Random Forest model compared to a cubist approach? 

"ökologischer Plausibilitätscheck"


# Methods

```{r load-packages, include=FALSE}

library(glcm)
library(raster) # required for dissimilarity
library(randomForest)
library(caret) # for k-fold cross-validation
library(sf)
library(gstat)
library(terra)
library(tmap)
library(stars)
library(devtools)
library(geodata)
library(spatstat) # for computing distance between sampling points
library(ggplot2)
library(tidyverse)

```

## Study area

The prediction of soil C:N was conducted for Germany. The shapefile of the administrative boundaries was obtained from `rnaturalearth` [@rnaturalearth].

```{r load-germany-mask}

library(rnaturalearth)
library(devtools)
devtools::install_github("ropensci/rnaturalearthhires", quiet = TRUE)

# get Germany boundary polygon
germany <- ne_countries(scale = 10, country = "Germany", returnclass = "sf") %>%
  select(geometry)

# get bounding box
st_bbox(germany)

# reproject to UTM 32N
crs <- 'EPSG:25832'
germany_4326 <- germany
germany <- st_transform(germany, crs)
```

## Data

## LUCAS soil survey

```{r load-lucas-data}

# load LUCAS topsoil data
lucas_eu <- st_read('data/LUCAS2015_topsoildata_20200323/LUCAS_Topsoil_2015_20200323-shapefile/LUCAS_Topsoil_2015_20200323.shp')

# reproject
lucas_eu <- st_transform(lucas_eu, crs)

# subset to Germany
lucas <- st_filter(lucas_eu, germany)

```
The measurements of soil properties were obtained from the LUCAS topsoil survey 2015. In Germany, it consisted of `r nrow(lucas)` samples. This data was used to calculate the C:N ratios by dividing soil organic carbon (SOC) by total nitrogen (N).

```{r calculate-cn-ratio}

# first make columns numeric, then add new column with ratio
lucas <- lucas %>%
  mutate(N = if_else(N == "< LOD", 0, as.numeric(N)),
         OC = if_else(OC == "< LOD", 0, as.numeric(OC)),
         cn_ratio = OC / N)

ggplot(lucas, aes(cn_ratio)) +
  geom_histogram(binwidth = 0.5) 
```

```{r remove-nans-and-outliers}

# copy before cleaning
lucas_init <- lucas

# summary statistics
cn_mean <- mean(lucas$cn_ratio)
cn_median <- median(lucas$cn_ratio)
cn_sd <- sd(lucas$cn_ratio)

iqr <- IQR(lucas$cn_ratio)
first_quantile <- quantile(lucas$cn_ratio)['25%']
third_quantile <- quantile(lucas$cn_ratio)['75%']

# remove nan and outliers (> IQR +- 1.5 * IQR)
lucas <- lucas %>%
  drop_na(cn_ratio) %>%
  mutate(cn_ratio_no_outlier = between(cn_ratio, first_quantile-1.5*iqr, third_quantile+1.5*iqr))

ggplot(lucas, aes(cn_ratio, fill=cn_ratio_no_outlier)) +
  geom_histogram(binwidth = 0.5) 
```
but doesn't really make sense because then we limit the prediction range of RF

```{r plot-lucas-points-germany}

tmap_mode("view")
tm_basemap('Esri.WorldImagery') +
  tm_shape(germany) +
  tm_lines(col="black", lwd = 3) +
  tm_shape(lucas) +
  tm_dots(fill="cn_ratio",
          size = 0.5,
          fill.scale = tm_scale_continuous(values = "cividis",
                                           limits = c(0, 20)))
```

### Distance between sampling points

```{r}

# prepare point pattern with projection (required for distance)
lucas_ppp <- as.ppp(lucas)

# find distance to nearest neighbor for each point
distances <- nndist(lucas_ppp)

# summary statistics
print(paste('Total number of points:', length(lucas$cn_ratio)))
print(paste('Mean distance:', round(mean(distances)/1000,2), 'km'))
print(paste('Median distance:', round(median(distances)/1000,2), 'km'))
print(paste('Standard deviation distance:', round(sd(distances)/1000,2), 'km'))
print(paste('Minimum distance:', round(min(distances)/1000,2), 'km'))
print(paste('Maximum distance:', round(max(distances)/1000,2), 'km'))

```

## Covariates

The selection of environmental predictors was based on those proposed by @khosravani_soil_2026 who had identified them according to expert knowledge. We slightly modified them when not avoidable due to lacking documentation in the paper or data availability. An overview of the used predictors can be found in TableXXX and in the following we describe all predictors more in detail.


| Abbreviation | Full name | Unit | Source |
|----|----|----|----|
| ann_temp | Mean annual temperature |  |  |
| ann_precip |  |  |  |
| slope | Slope |  | https://zenodo.org/records/1447210 |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |

: Covariates used in C:N prediction, as proposed by Khosravani et al. (2025)


elevation, Slope and aspect
- Used DEM from Hengl / Yamazaki et al (https://zenodo.org/records/1447210).
- used terra [@terra] for calculating slope and aspect

Temperature and precipitation
-   loaded from WorldClim (XXX)

Latitude
extracted latitude from rasters

```{r load-covariate-data-1}

# load DEM raster
dem_global <- rast('data/covariates/elevation/dtm_elevation_merit.dem_m_250m_s0..0cm_2017_v1.0.tif') %>%
  crop(germany_4326)

# calculate slope 
slope <- terrain(dem, v = "slope", unit = "degrees")

# calculate aspect
aspect <- terrain(dem, v = "aspect", unit = "degrees")

# WorldClim (temperature and precipitation)
temp <- worldclim_country(country='germany', var='tavg', path='data/covariates/worldclim', res=0.5)
precip <- worldclim_country(country='germany', var='prec', path='data/covariates/worldclim', res=0.5)

temp_annual <- mean(temp) # mean annual temperature
precip_annual <- sum(precip) # total annual precipitation

# latitude
latitude <- rast(list(latitude = init(dem, "y")))

```


Land cover (dissimilarity)

link in paper was not accessible

instead https://zenodo.org/records/3939038

but is dissimilarity really more "aussagekraeftig" than the categorical LC type?
maybe try if adding LC type would increase accuracy?

```{r lc-dissimilarity}

# set option to redo the analysis (take a few minutes) or load existing raster
redo_lc_dissim <- FALSE

# set path
lc_dissim_path <- "data/covariates/lc_dissim.tif"

# load land cover data and crop to Germany
land_cover_global <- rast('data/covariates/land-cover/PROBAV_LC100_global_v3.0.1_2015-base_Discrete-Classification-map_EPSG-4326.tif')
land_cover <- crop(land_cover_global, germany_4326)

# check if file exists AND redo flag is FALSE --> just load existing raster
if (file.exists(lc_dissim_path) && !redo_lc_dissim) {
  
  lc_dissim <- rast(lc_dissim_path)
  
} else {
  
  # transform to RasterLayer (from raster package) - required for glcm
  land_cover_rast <- raster(land_cover)
  
  # dissimilarity (from ChatGPT)
  dissim_rast <- glcm(
    land_cover_rast,
    window = c(3,3), # take only the 8 pixels forming the square directly around current pixel
    shift = list(c(1,0), c(0,1), c(1,1), c(-1,1)), # average 4 directions (up, down, left, right)
    statistics = "dissimilarity"
  )
  
  # convert back to terra
  lc_dissim <- rast(dissim_rast)
  
  # save
  writeRaster(lc_dissim, lc_dissim_path, overwrite=TRUE)
}

```


Clay and silt content
very poor documentation in paper!
there it says from ISRIC
eventually downloaded from SoilGrids (250m resolution)
- clay content (g kg-1)
- in topsoil ( 0 to 5 cm)

pH
soilgrids

Soil water content (SWC)
Vol. water content at -33 kPa from SoilGrids (at field capacity)
soilgrids

Soil respiration allegedly from ISRIC, but only found link there to https://www.earthdata.nasa.gov/data/catalog/ornl-cloud-soilresp-heterotrophicresp-1928-1
in supplementary materials in g CO2 per m2, but in NASA data g C per m2 per y


HFP from https://datadryad.org/dataset/doi:10.5061/dryad.3tx95x6d9

NDVI
- not specified in paper how exactly calculated (besides GEE)
- We used a 3-year growing season median (to capture long-term trends and avoid overrepresentation of weather influences; since we see C:N as a long-term property rather than a momentary snapshot)
- see GEE script in repository

```{r load-covariate-data-2}

clay <- rast('data/covariates/clay_0-5.tif') %>%
  crop(germany_4326)
silt <- rast('data/covariates/silt_0-5.tif') %>%
  crop(germany_4326)

# pH
ph <- rast('data/covariates/ph_0-5.tif') %>%
  crop(germany_4326)

# soil water content (SWC)
swc <- rast('data/covariates/swc.tif') %>%
  crop(germany_4326)

# soil respiration
s_resp <- rast('data/SoilResp_HeterotrophicResp_1928/data/soil_Rs_mean.tif') %>%
  crop(germany_4326)

# human footprint (HFP)
hfp <- rast('data/covariates/Human_footprint_maps/hfp2013_merisINT.tif') %>%
  crop(st_transform(germany, 'EPSG:9001'))

# Normalized Difference Vegetation Index (NDVI)
ndvi <- rast('data/covariates/NDVI_Germany_2013_2015.tif')

```



Once we had obtained the raster data for all predictors, we projected them to the same CRS (EPSG 25832, UTM 32N) and resolution. The reference raster chosen for this alignment was the slope since it already had the target resolution of 250 m.

From the resulting raster stack, we removed pixels corresponding to unsuitable land cover types (water bodies, snow, built-up land).

```{r prepare-raster-stack}

# same CRS and aligned

# set slope raster as template to align all other rasters with (because it has target resolution 250m)
slope <-  project(slope, crs, method='bilinear') 

# add to stack
predictor_stack <- c(slope)

# realign all remaining raster layers
predictors <- list(dem, aspect, temp_annual, precip_annual, latitude, lc_dissim, clay, silt, ph, swc, s_resp, hfp, ndvi, land_cover)
for (predictor in predictors) {
  predictor <- project(predictor, slope, method="bilinear")
  predictor_stack <- append(predictor_stack, predictor)
}

# adjust names
names(predictor_stack) <- c('elevation', 'slope', 'aspect', 'mean_temp', 'total_precip', 'latitude', 'lc_dissim', 'clay', 'silt', 'ph', 'swc', 's_resp', 'hfp', 'ndvi', 'lc_class')

# mask with Germany (extract exact shape)
predictor_stack <- mask(predictor_stack, germany)

# mask out certain land cover types
unsuitable_lc_classes <- c(
  0, # no input data available
  50, # urban/built-up
  70, # snow and ice
  80, # permanent water bodies
  200 # open sea
)
suitable_lc <- app(predictor_stack$lc_class, fun=function(x){ x[x %in% unsuitable_lc_classes] <- NA; return(x)} ) # set pixels corresponding to those LC types to NA
predictor_stack <- mask(predictor_stack, suitable_lc) # set to NA in whole stack

plot(predictor_stack)

```


## Random Forest model


```{r split-train-test}

set.seed(123)

# extract data for each point from raster stack
predictor_values <- terra::extract(predictor_stack, lucas)[,-1] # remove ID

# add extracted values to C:N ratio data
data <- cbind(lucas[c('Point_ID', 'cn_ratio')], predictor_values) %>%
  st_drop_geometry()

# randomly select points by proportion
train_data <- data %>%
  slice_sample(prop = 0.8) 
  
train_ids <- train_data$Point_ID

# use remaining rows for test
test_data <- data %>%
  filter_out(Point_ID %in% train_ids) %>% # keep only IDs NOT in training data
  select(!Point_ID)

```



```{r train-random-forest-model}

set.seed(123)

# remove Point_ID
train_data <- select(train_data, !Point_ID)

# train model
rf_model <- randomForest(
  cn_ratio ~ .,      # C:N as function of all other columns (predictors)
  data = train_data,
  ntree = 500,       # number of trees
  importance = TRUE  # calculate variable importance scores
)

rf_model

```

```{r test-model}

train_pred <- predict(
  rf_model,
  newdata = train_data,
)

plot(train_data$cn_ratio, train_pred)
abline(0, 1, col='red')
```

```{r spatial-prediction}

cn_map <- terra::predict(
  object = predictor_stack, # what to predict from
  model  = rf_model
)

names(cn_map) <- c("Mean predicted C:N ratio")

plot(cn_map)

```

The reference paper by @khosravani_soil_2026 only used k-fold cross-validation for hyperparameter tuning and apparently calculated training and test based on a single split. We decided to implement a k-fold cross-validation (CV) for calculating the accuracy metrics as well, to obtain more robust estimates. In k-fold CV, the dataset is split into k folds, i.e., approximately equal partitions. 
We implemented this k-fold CV using the `caret` [@caret} package in R, setting k to 10.

Both the test set accuracy as well as the metrics obtained via CV evaluate the model performance when predicting from unseen data. We therefore decided to focus on the comparison of these two, leaving out training performance since it is much less directly relevant for assessing the model. 

```{r prepare-cross-validation}

set.seed(123)

# 10-fold CV (split model into 10 folds, run 10 times, each time, 1 fold is test while remaining 9 are used for training)
train_control <- trainControl(method = "cv", 
                              number = 10,
                              savePredictions = "final"
                              )

rfModel <- train(cn_ratio ~ ., # C:N ratio as a function of all predictors
                 data = select(data, !Point_ID), 
                 method = "rf", # random forest
                 trControl = train_control, # implement CV as defined above
                 importance = TRUE) # compute permutation importance

rfModel
```

Values refer to mean values across 10 folds

rfModel$pred 
Each row is:

A prediction for one observation

From the fold where it was in the validation set

So these are pure out-of-fold predictions.

```{r}
best_mtry <- rfModel$bestTune$mtry

cv_preds <- rfModel$pred
cv_preds <- cv_preds[cv_preds$mtry == best_mtry, ]

RMSE <- sqrt(mean((cv_preds$obs - cv_preds$pred)^2))
R2 <- cor(cv_preds$obs, cv_preds$pred)^2
Bias <- mean(cv_preds$pred - cv_preds$obs)

x <- cv_preds$pred
y <- cv_preds$obs

rho <- cor(x, y)
sd_x <- sd(x)
sd_y <- sd(y)
mean_x <- mean(x)
mean_y <- mean(y)

CCC_value <- (2 * rho * sd_x * sd_y) /
  (sd_x^2 + sd_y^2 + (mean_x - mean_y)^2)
```

## Cubist model

```{r}
library(Cubist)

train_pred <- select(train_data, !cn_ratio)
train_resp <- train_data$cn_ratio

model_tree <- cubist(x = train_pred, y = train_resp)
model_tree

summary(model_tree)

```

```{r predict-cubist}

cn_map_cubist <- terra::predict(
  object = predictor_stack, # what to predict from
  model  = model_tree
)

names(cn_map_cubist) <- c("Mean predicted C:N ratio (cubist)")

cn_map_cubist <- mask(cn_map_cubist, germany)

plot(cn_map_cubist)

```

```{r difference-map}

diff_map <- cn_map - cn_map_cubist
plot(diff_map)

# Define breakpoints and colors
breaks <- c(minmax(diff_map)[1], -2, 2, minmax(diff_map)[2])

colors <- c("red",        # below -5
            "lightgrey",  # between -5 and 5
            "blue")       # above 5

# Plot with custom colors
plot(diff_map, breaks = breaks, col = colors, legend = TRUE)
```

## Model evaluation

-   coefficient of determination (R2)
-   concordance correlation coefficient (CCC)
-   root mean squared error (RMSE)
-   Bias

In the reference paper, it is poorly documented from which training and testing sets the accuracy metrics were obtained, so we followed the approach of @zhou_prediction_2021 and calculated them as the mean from 10-fold cross-validation to have more robust estimates.

```{r calculate-accuracy-measures}


```

### Uncertainty map

via bootstrapping --> mean and SD map

e.g., https://www.sciencedirect.com/science/article/pii/S0048969720361908 created 100 maps

```{r}

```


## Ecological plausibility

- Natura 2000 shapefile obtained from https://www.eea.europa.eu/en/datahub/datahubitem-view/6fc8ad2d-195d-40f4-bdec-576e7d1268e4

# Results

## Exploratory data analysis

### C:N ratio

#### Spatial autocorrelation

```{r c-n-experimental-variogram}

# use projection (in m)
lucas_proj <- st_transform(lucas, 'EPSG:25832')

# make gstat object
gstat_cn <- gstat(formula = cn_ratio ~ 1, data = lucas_proj)

vg_cn <- variogram(gstat_cn, cloud=F, boundaries=seq(3000, 30000, 3000))
plot(vg_cn, plot.nu = TRUE)

vg_cn <- variogram(gstat_cn, cloud=F, boundaries=seq(2000, 10000, 2000))
plot(vg_cn, plot.nu = TRUE)

```
--> spatial autocorrelation only to distance of ~7.5km and only small fraction of points have other points within this distance (especially when trying variogram with smaller cloud classes), so geostatistical approaches such as kriging not ideal

### Covariates

#### Collinearity

```{r}
terra::layerCor(predictor_stack, fun='cor')
```

pH and SWC had a high (>0.9) correlation.

## Variable importance

```{r plot-variable-importance}
varImpPlot(rf_model, main = "Predictor Importance")

varImpPlot(rfModel$finalModel)
```

## Evaluation

### Summary statistics of predicted C:N by land cover class

# Discussion



## Environmental predictors

Identical to the results by @khosravani_soil_2026, pH was also by far the most important predictor in both our models. 

A notable difference was the importance of NDVI. While @khosravani_soil_2026 had a relative importance of <5% for this predictor, in our models, it was the second most important covariate. 
However, it needs to be noted that it was not clear how @khosravani_soil_2026 had computed the NDVI map. We wanted to capture it as a long-term indicator of productivity by using a 3-year growing-season median, which seemed to work well.

*But does NDVI affect the C:N ratio directly or is it more likely to be a proxy for some other property?*

## Limitations

-   WorldClim data from 1970 to 2000 - climate change might already have had an effect by 2018
- 


# AI documentation

## Preparation phase

-   get an overview of possible research directions
-   refine search terms

# References