---
title: "Digital soil mapping of C:N ratio across Germany"
author: " & Nina Preußler"
format:
  html:
    self-contained: true 
execute: # set global chunk options
  echo: true # show code 
  warning: false # don't show warnings
  error: false # only render if no errors arise
bibliography: references.bib 
csl: https://www.zotero.org/styles/apa # citation style
editor: visual
knitr:
  opts_knit:
    root.dir: "~/Documents/00_ESDSRS/semester1/3_soil/soil-N-project" # working directory
---

# Title

Comparison of Random Forest and cubist machine learning for spatial prediction of soil C:N stoichiometry across Germany accuracy measures, spatial patterns, variable importance

# Introduction

## Why is C:N important?

-   What does it tell you about the soil?

## How is C:N changing?

-   What processes and factors affect it?
-   EU/Germany trends

## Why DSM?

-   existing maps for EU/Germany
    -   method
    -   data
    -   quality
    -   challenges

## Fragestellung

-   methodische Gap: keine Random Forest prediction in Deutschland

Schlagwörter: - Random Forest - C:N ratio Stöchiometrie - systematischer Vergleich - accuracy (Genauigkeit?) - spatial patterns (z.B. Boxplot nach land cover, elevation; visueller Eindruck, hotspots) - Variablen (importance, partial dependence)

How does spatial prediction of soil C:N stoichiometry across Germany differ when using a Random Forest model compared to a cubist approach?

"ökologischer Plausibilitätscheck"

# Methods

```{r load-packages}
#| include: false

library(glcm)
library(Cubist)
library(raster) # required for dissimilarity
library(randomForest)
library(caret) # for k-fold cross-validation
library(sf)
library(gstat)
library(terra)
library(tmap)
library(devtools)
library(geodata)
library(spatstat) # for computing distance between sampling points
library(ggplot2)
library(patchwork)
library(tidyverse)

options(renv.config.startup.quiet = TRUE)
```

## Study area

The prediction of soil C:N was conducted for Germany. The shapefile of the administrative boundaries was obtained from `rnaturalearth` [@rnaturalearth].

```{r load-germany-mask}

library(rnaturalearth)
library(devtools)
devtools::install_github("ropensci/rnaturalearthhires", quiet = TRUE)

# get Germany boundary polygon
germany <- ne_countries(scale = 10, country = "Germany", returnclass = "sf") %>%
  select(geometry)

# get bounding box
st_bbox(germany)

# reproject to UTM 32N
crs <- 'EPSG:25832'
germany_4326 <- germany
germany <- st_transform(germany, crs)
```

## Data

### LUCAS soil survey

```{r load-lucas-data}

# load LUCAS topsoil data
lucas_eu <- st_read('data/LUCAS2015_topsoildata_20200323/LUCAS_Topsoil_2015_20200323-shapefile/LUCAS_Topsoil_2015_20200323.shp')

# reproject
lucas_eu <- st_transform(lucas_eu, crs)

# subset to Germany
lucas <- st_filter(lucas_eu, germany)

```

The measurements of soil properties were obtained from the LUCAS topsoil survey 2015. In Germany, it consisted of `r nrow(lucas)` samples. This data was used to calculate the C:N ratios by dividing soil organic carbon (SOC) by total nitrogen (N).

```{r calculate-cn-ratio}
#| fig-cap: "Distribution of C:N ratio values from LUCAS 2015 samples in Germany."
#| fig-height: 3
#| fig-dpi: 200

# first make columns numeric, then add new column with ratio
lucas <- lucas %>%
  mutate(N = if_else(N == "< LOD", 0, as.numeric(N)),
         OC = if_else(OC == "< LOD", 0, as.numeric(OC)),
         cn_ratio = OC / N)

ggplot(lucas, aes(cn_ratio)) +
  geom_histogram(binwidth = 0.5) +
  labs(x = "C:N ratio", y = "number of samples")
```

```{r remove-nans-and-outliers}
#| eval: false # don't execute
#| include: false

# copy before cleaning
lucas_init <- lucas

# summary statistics
cn_mean <- mean(lucas$cn_ratio)
cn_median <- median(lucas$cn_ratio)
cn_sd <- sd(lucas$cn_ratio)

iqr <- IQR(lucas$cn_ratio)
first_quantile <- quantile(lucas$cn_ratio)['25%']
third_quantile <- quantile(lucas$cn_ratio)['75%']

# remove nan and outliers (> IQR +- 1.5 * IQR)
lucas <- lucas %>%
  drop_na(cn_ratio) %>%
  mutate(cn_ratio_no_outlier = between(cn_ratio, first_quantile-1.5*iqr, third_quantile+1.5*iqr))

ggplot(lucas, aes(cn_ratio, fill=cn_ratio_no_outlier)) +
  geom_histogram(binwidth = 0.5) 
```

There was one notable outlier at a ratio of almost 40, but we did not remove it because it helps increases the prediction range of the random forest algorithm. The extreme value most likely did not arise from measuring error since a comparable value can also be seen in the later LUCAS 2018 survey.

```{r plot-lucas-points-germany}
#| include: false
#| eval: false

tmap_mode("view")
tm_basemap('Esri.WorldImagery') +
  tm_shape(germany) +
  tm_lines(col="black", lwd = 3) +
  tm_shape(lucas) +
  tm_dots(fill="cn_ratio",
          size = 0.5,
          fill.scale = tm_scale_continuous(values = "cividis",
                                           limits = c(0, 20)))
```

### Distance between sampling points

```{r}

# prepare point pattern with projection (required for distance)
lucas_ppp <- as.ppp(lucas)

# find distance to nearest neighbor for each point
distances <- nndist(lucas_ppp)

# summary statistics
print(paste('Total number of points:', length(lucas$cn_ratio)))
print(paste('Mean distance:', round(mean(distances)/1000,2), 'km'))
print(paste('Median distance:', round(median(distances)/1000,2), 'km'))
print(paste('Standard deviation distance:', round(sd(distances)/1000,2), 'km'))
print(paste('Minimum distance:', round(min(distances)/1000,2), 'km'))
print(paste('Maximum distance:', round(max(distances)/1000,2), 'km'))

```

## Covariates

The selection of environmental predictors was based on those proposed by @khosravani_soil_2026 who had identified them according to expert knowledge. We slightly modified them when not avoidable due to lacking documentation in the paper or data availability. An overview of the used predictors can be found in TableXXX and in the following we describe all predictors more in detail.

| Abbreviation | Full name | Unit | Source |
|----|----|----|----|
| ann_temp | Mean annual temperature |  |  |
| ann_precip |  |  |  |
| slope | Slope |  | https://zenodo.org/records/1447210 |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |
|  |  |  |  |

: Covariates used in C:N prediction, as proposed by Khosravani et al. (2025)

elevation, Slope and aspect - Used DEM from Hengl / Yamazaki et al (https://zenodo.org/records/1447210). - used terra [@terra] for calculating slope and aspect

Temperature and precipitation - loaded from WorldClim (XXX)

Latitude extracted latitude from rasters

```{r load-covariate-data-1}

# load DEM raster
dem <- rast('data/covariates/elevation/dtm_elevation_merit.dem_m_250m_s0..0cm_2017_v1.0.tif') %>%
  crop(germany_4326)

# calculate slope 
slope <- terrain(dem, v = "slope", unit = "degrees")

# calculate aspect
aspect <- terrain(dem, v = "aspect", unit = "degrees")

# WorldClim (temperature and precipitation)
temp <- worldclim_country(country='germany', var='tavg', path='data/covariates/worldclim', res=0.5)
precip <- worldclim_country(country='germany', var='prec', path='data/covariates/worldclim', res=0.5)

temp_annual <- mean(temp) # mean annual temperature
precip_annual <- sum(precip) # total annual precipitation

# latitude
latitude <- rast(list(latitude = init(dem, "y")))

```

Land cover (dissimilarity)

link in paper was not accessible

instead https://zenodo.org/records/3939038

but is dissimilarity really more "aussagekraeftig" than the categorical LC type? maybe try if adding LC type would increase accuracy?

```{r lc-and-dissimilarity}

# set option to redo the analysis (take a few minutes) or load existing raster
redo_lc_dissim <- FALSE

# set path
lc_dissim_path <- "data/covariates/lc_dissim.tif"

# load land cover data and crop to Germany
land_cover <- rast('data/covariates/land-cover/PROBAV_LC100_global_v3.0.1_2015-base_Discrete-Classification-map_EPSG-4326.tif') %>%
  crop(germany_4326)

# obtain LC dissimilarity
# check if file exists AND redo flag is FALSE --> just load existing raster
if (file.exists(lc_dissim_path) && !redo_lc_dissim) {
  
  lc_dissim <- rast(lc_dissim_path)
  
} else {
  
  # transform to RasterLayer (from raster package) - required for glcm
  land_cover_rast <- raster(land_cover)
  
  # dissimilarity (from ChatGPT)
  dissim_rast <- glcm(
    land_cover_rast,
    window = c(5,5), # take only the 8 pixels forming the square directly around current pixel
    shift = list(c(1,0), c(0,1), c(1,1), c(-1,1)), # average 4 directions (up, down, left, right)
    statistics = "dissimilarity"
  )
  
  # convert back to terra
  lc_dissim <- rast(dissim_rast)
  
  # save
  writeRaster(lc_dissim, lc_dissim_path, overwrite=TRUE)
}

```

Clay and silt content very poor documentation in paper! there it says from ISRIC eventually downloaded from SoilGrids (250m resolution) - clay content (g kg-1) - in topsoil ( 0 to 5 cm)

pH soilgrids

Soil water content (SWC) Vol. water content at -33 kPa from SoilGrids (at field capacity) soilgrids

Soil respiration allegedly from ISRIC, but only found link there to https://www.earthdata.nasa.gov/data/catalog/ornl-cloud-soilresp-heterotrophicresp-1928-1 in supplementary materials in g CO2 per m2, but in NASA data g C per m2 per y

HFP from https://datadryad.org/dataset/doi:10.5061/dryad.3tx95x6d9

NDVI - not specified in paper how exactly calculated (besides GEE) - We used a 3-year growing season median (to capture long-term trends and avoid overrepresentation of weather influences; since we see C:N as a long-term property rather than a momentary snapshot) - see GEE script in repository

```{r load-covariate-data-2}

clay <- rast('data/covariates/clay_0-5.tif') %>%
  crop(germany_4326)
silt <- rast('data/covariates/silt_0-5.tif') %>%
  crop(germany_4326)

# pH
ph <- rast('data/covariates/ph_0-5.tif') %>%
  crop(germany_4326)

# soil water content (SWC)
swc <- rast('data/covariates/swc.tif') %>%
  crop(germany_4326)

# soil respiration
s_resp <- rast('data/SoilResp_HeterotrophicResp_1928/data/soil_Rs_mean.tif') %>%
  crop(germany_4326)

# human footprint (HFP)
hfp <- rast('data/covariates/Human_footprint_maps/hfp2013_merisINT.tif') 
hfp <- crop(hfp, st_transform(germany, crs(hfp)))

# Normalized Difference Vegetation Index (NDVI)
ndvi <- rast('data/covariates/NDVI_Germany_2013_2015.tif')

```

Once we had obtained the raster data for all predictors, we projected them to the same CRS (EPSG 25832, UTM 32N) and resolution. The reference raster chosen for this alignment was the slope since it already had the target resolution of 250 m.

From the resulting raster stack, we removed pixels corresponding to unsuitable land cover types (water bodies, snow, built-up land).

```{r prepare-raster-stack}

# same CRS and aligned

# set slope raster as template to align all other rasters with (because it has target resolution 250m)
slope <-  project(slope, crs, method='bilinear') 

# add to stack
predictor_stack <- c(slope)

# realign all remaining raster layers
predictors <- list(dem, aspect, temp_annual, precip_annual, latitude, lc_dissim, clay, silt, ph, swc, s_resp, hfp, ndvi)
for (predictor in predictors) {
  predictor <- project(predictor, slope, method="bilinear")
  predictor_stack <- append(predictor_stack, predictor)
}

# add land cover (categorical, therefore use different reprojection method)
land_cover_reproj <- project(land_cover, slope, method='mode') # majority voting
predictor_stack <- append(predictor_stack, land_cover_reproj)

# adjust names
names(predictor_stack) <- c('slope', 'elevation', 'aspect', 'mean_temp', 'total_precip', 'latitude', 'lc_dissim', 'clay', 'silt', 'ph', 'swc', 's_resp', 'hfp', 'ndvi', 'lc_class')

# mask with Germany (extract exact shape)
predictor_stack <- mask(predictor_stack, germany)

# mask out certain land cover types
unsuitable_lc_classes <- c(
  0, # no input data available
  50, # urban/built-up
  70, # snow and ice
  80, # permanent water bodies
  200 # open sea
)
suitable_lc <- app(predictor_stack$lc_class, fun=function(x){ x[x %in% unsuitable_lc_classes] <- NA; return(x)} ) # set pixels corresponding to those LC types to NA
predictor_stack <- mask(predictor_stack, suitable_lc) # set to NA in whole stack

# add dummy variables encoding major land cover types
forest_codes <- c(111,112,113,114,115,116,121,122,123,124,125,126)
forest <- predictor_stack$lc_class %in% forest_codes  # returns binary raster (True/False)

grassland <- predictor_stack$lc_class == 30 
wetland <- predictor_stack$lc_class == 90 
cropland <- predictor_stack$lc_class == 40 

# add to predictor stack
predictor_stack <- c(predictor_stack, forest*1, grassland*1, wetland*1, cropland*1)
names(predictor_stack) <- c(names(predictor_stack)[1:(length(names(predictor_stack))-4)], 'forest', 'grassland', 'wetland', 'cropland')

plot(predictor_stack)

```

## Random Forest model


```{r prepare-model-data}

# extract data for each point from raster stack
predictor_values <- terra::extract(predictor_stack, lucas)[,-1] # remove ID

# combine with C:N data and remove unnecessary colums
data <- cbind(lucas['cn_ratio'], predictor_values) %>%
  drop_na() %>%
  st_drop_geometry() 

nrow(data)
```

The reference paper by @khosravani_soil_2026 only used k-fold cross-validation for hyperparameter tuning and apparently calculated training and test based on a single split. We decided to implement a k-fold cross-validation (CV) for calculating the accuracy metrics as well, to obtain more robust estimates. In k-fold CV, the dataset is split into k folds, i.e., approximately equal partitions. We implemented this k-fold CV using the `caret` \[@caret} package in R, setting k to 10. Caret automatically conducted hyperparameter tuning for the parameter `mtry`, i.e., the number of predictors used at each split.

Both the test set accuracy as well as the metrics obtained via CV evaluate the model performance when predicting from unseen data. We therefore decided to focus on the comparison of these two, leaving out training performance since it is much less directly relevant for assessing the model.

CV also used for hyperparameter tuning, specifically to pick the optimal number for `mtry` which denotes the number of predictors to use at each split.

```{r rf-cross-validation}

set.seed(123)

# 10-fold CV (split model into 10 folds, run 10 times, each time, 1 fold is test while remaining 9 are used for training)
train_control <- trainControl(method = "cv", 
                              number = 10,
                              savePredictions = "final"
                              )


rfModel <- train(cn_ratio ~ ., # C:N ratio as a function of all predictors
                 data = data, 
                 method = "rf", # random forest
                 tuneLength = 5,
                 ntree = 500,
                 trControl = train_control, # implement CV as defined above
                 importance = TRUE) # compute permutation importance

rfModel
```
Hyperparameter tuning via `caret` indicated that `mtry` `r rfModel$bestTune$mtry` had the lowest RMSE.


## Cubist model

While we originally wanted to compare our results from the Random Forest Regression with the Cubist predicitons from @khosravani_soil_2026, we were not able to accurately replicate their methodology (particularly the covariate layer acquisition and preprocessing) due to gaps in documentation. 

We therefore implemented a cubist model ourselves, that is trained in the same way (10-fold CV) as the RF.

```{r cubist-cross-validation}

library(Cubist)

set.seed(123)

cubModel <- train(cn_ratio ~ ., # C:N ratio as a function of all predictors
                 data = data, 
                 method = "cubist", 
                 trControl = train_control, # implement CV as defined above for Random Forest
                 importance = TRUE) # compute permutation importance

cubModel
```

`r cubModel$bestTune$committees`  committees and `r cubModel$bestTune$neighbors`  neighbors led to the lowest RMSE.

## Variable importance

calculated from the best-fit model trained on all data after CV




## Model performance

Accuracy metrics

-   coefficient of determination (R2)
-   concordance correlation coefficient (CCC)
-   root mean squared error (RMSE)
-   Bias

In the reference paper, it is poorly documented from which training and testing sets the accuracy metrics were obtained, so we followed the approach of @zhou_prediction_2021 and calculated them as the mean from 10-fold cross-validation to have more robust estimates.

```{r calculate-accuracy-measures}

# extract observations, predictions, and fold columns from best-fit model
rf_preds <- rfModel$pred %>%
  select(pred, obs, Resample)

# calculate accuracy metrics per fold
rf_preds_accuracy <- rf_preds %>%
  group_by(Resample) %>%
  summarise(
    RMSE = sqrt(mean((obs - pred)^2)),
    R2 = 1 - ((mean((obs - pred)^2))/(mean((obs - mean(rf_preds$obs))^2))),
    Bias = mean(obs - pred),
    CCC = (2 * cor(rf_preds$obs, rf_preds$pred) * sd(rf_preds$obs) * sd(rf_preds$pred)) /
  (var(rf_preds$obs) + var(rf_preds$pred) + (mean(obs) - mean(pred))^2)
)

# same for cubist
cubist_preds <- cubModel$pred %>%
  select(pred, obs, Resample)

cubist_preds_accuracy <- cubist_preds %>%
  group_by(Resample) %>%
  summarise(
    RMSE = sqrt(mean((obs - pred)^2)),
    R2 = 1 - ((mean((obs - pred)^2))/(mean((obs - mean(cubist_preds$obs))^2))),
    Bias = mean(obs - pred),
    CCC = (2 * cor(cubist_preds$obs, cubist_preds$pred) * sd(cubist_preds$obs) * sd(cubist_preds$pred)) /
  (var(cubist_preds$obs) + var(cubist_preds$pred) + (mean(obs) - mean(pred))^2)
)

# combine in one df
rf_preds_accuracy$model <- "Random Forest"
cubist_preds_accuracy$model <- "Cubist"
accuracies <- rbind(rf_preds_accuracy, cubist_preds_accuracy)

# average within each model across folds
accuracies_summarized <- accuracies %>%
  group_by(model) %>%
  summarise(across(everything(), mean)) %>%
  select(!Resample)
  
```

## Mapping

To obtain mean and uncertainty maps, we used a bootstrapping approach as @khosravani_soil_2026 did. Due to computational constraints, we used 30 iteration for this process. In each iteration, a cubist and a random forest model were trained on a bootstrapped sample of the data and then used to make spatial predictions based on the predictor stack. From the resulting 30 maps per model, we calculated the mean predicted C:N as well as uncertainty map based on the standard deviation. To investigate possible spatial patterns between the RF and Cubist predictions, we also computed a difference map by substracting the mean Cubist from the mean RF map.

```{r bootstrapping}

# TAKES LONG! only redo if something about data changed
redo_bootstrap <- FALSE

# set number of bootstrapping runs
b_max = 5

n = nrow(data)

# downsample raster stack for testing
predictor_stack_coarse <- terra::aggregate(predictor_stack, fact = 3)

if (redo_bootstrap) {
  
  for (b in seq(1,b_max)) {
    
    # progress report
    cat("Bootstrap iteration:", b, "of", b_max, "\n")
    
    # get indexes via bootstrap sampling (with replacement)
    idxs <- sample(n, size=n, replace = TRUE)
    
    # subset data
    boot_data <- data[idxs,]
    
    # train models (with best hyperparameter tuning values) on it
    rf_model <- randomForest(
      cn_ratio ~ .,      # C:N as function of all other columns (predictors)
      data = boot_data,
      mtry = rfModel$bestTune$mtry,
      ntree = 500       # number of trees
      )
    cubist_model <- cubist(
      x = select(boot_data, !cn_ratio), 
      y = boot_data$cn_ratio,
      committees = cubModel$bestTune$committees
      )
    
    # spatial prediction
    map_rf <- terra::predict(
      object = predictor_stack_coarse, # what to predict from
      model  = rf_model,
      cores=4, # parallel processing
      cpkgs = "randomForest", # package needed to run prediction
      filename = paste0('data/bootstrap_prediction_layers/rf_', b, '.tif' ), # write to disk instead of keeping in memory
      overwrite = TRUE
      )
    map_cubist <- terra::predict(
      object = predictor_stack_coarse, 
      model  = cubist_model,
      cores=4,
      cpkgs = "Cubist", 
      filename = paste0('data/bootstrap_prediction_layers/cub_', b, '.tif' ), 
      overwrite = TRUE
    )
  }
}

```


```{r}

# load bootstrapping map predictions
path <- 'data/bootstrap_prediction_layers/'
rf_boot_stack <- rast(paste0(path, list.files(path, pattern='rf*')))
cub_boot_stack <- rast(paste0(path, list.files(path, pattern='cub*')))

# mask cubist stack with germany
cub_boot_stack <- mask(cub_boot_stack, germany)

# compute mean and standard deviation across layers
rf_mean_map <- app(rf_boot_stack, mean)
rf_sd_map   <- app(rf_boot_stack, sd)

cub_mean_map <- app(cub_boot_stack, mean)
cub_sd_map   <- app(cub_boot_stack, sd)

# difference map
diff_map <- rf_mean_map - cub_mean_map

```

e.g., https://www.sciencedirect.com/science/article/pii/S0048969720361908 created 100 maps


## Ecological plausibility

To illustrate a possible use case for C:N maps, we tested whether 

For this purpose, we used a shapefile of Natura 2000 areas obtained from https://www.eea.europa.eu/en/datahub/datahubitem-view/6fc8ad2d-195d-40f4-bdec-576e7d1268e4. 

```{r load-natura-2000}

natura <- st_read('data/SHP/Natura2000_end2024_epsg3035.shp') %>%
  st_transform(crs) %>%
  st_intersection(germany)

# mask with shapefile to get distribution within and outside pro
rf_mean_map_natura <- mask(rf_mean_map, natura)
rf_mean_map_not_natura <- mask(rf_mean_map, natura, inverse=TRUE)

cub_mean_map_natura <- mask(cub_mean_map, natura)
cub_mean_map_not_natura <- mask(cub_mean_map, natura, inverse=TRUE)
```

# Results

## Model performance

- models had very similar performance
- RF slightly higher R2 and lower RMSE, see @accuracy-table
- RF tends to overestimate values (slightly negative bias), especially lower observed values
- cubist tends to underestimate (positive bias)
- cubist had higher CCC, also visible in @plot-obs-vs-pred
- 

```{r accuracy-table}
accuracies_summarized
```


```{r plot-obs-vs-pred}

# two-panel plot with RF and cubist

# Bin size control + color palette
ggplot(rf_preds, aes(obs, pred) ) +
  geom_hex(bins = 70) +
  scale_fill_continuous(type = "viridis") +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  coord_equal() 

ggplot(cubist_preds, aes(obs, pred) ) +
  geom_hex(bins = 70) +
  scale_fill_continuous(type = "viridis") +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  coord_equal() #+
  # xlim(5, 20) +
  # ylim(5,20)

```


## Maps

```{r mean-and-uncertainty-maps}
plot(rf_mean_map, main='Mean prediction Random Forest')
plot(rf_sd_map, main='Uncertainty map Random Forest')

plot(cub_mean_map, main='Mean prediction Cubist')
plot(cub_sd_map, main='Uncertainty map Cubist')
```


```{r plot-difference-map}

# Define breakpoints and colors
breaks <- c(minmax(diff_map)[1], -1, 1, minmax(diff_map)[2])

colors <- c("red",        # below -2
            "lightgrey",  # between -2 and 2
            "blue")       # above 2

# Plot with custom colors
plot(diff_map, breaks = breaks, col = colors, legend = TRUE)

```


```{r correlation-rf-cubist}
cor(arrange(rfModel$pred, rowIndex)['pred'], arrange(cubModel$pred, rowIndex)['pred'])
```


## Variable importance

```{r plot-variable-importance}
varImpPlot(rfModel$finalModel, main = "Predictor Importance Random Forest")
```

```{r variable-importance}
varImp(rfModel)

varImp(cubModel)
```



## Evaluation

### Summary statistics of predicted C:N by land cover class


### Ecological plausibility

```{r compare-natura-boxplot}

rf_natura <- values(rf_mean_map_natura)
rf_not_natura <- values(rf_mean_map_not_natura)
cub_natura <- values(cub_mean_map_natura)
cub_not_natura <- values(cub_mean_map_not_natura)

boxplot(cbind(rf_natura, rf_not_natura, cub_natura, cub_not_natura))
```


# Discussion

## Environmental predictors

Identical to the results by @khosravani_soil_2026, pH was also by far the most important predictor in both our models.

A notable difference was the importance of NDVI. While @khosravani_soil_2026 had a relative importance of \<5% for this predictor, in our models, it was the second most important covariate. However, it needs to be noted that it was not clear how @khosravani_soil_2026 had computed the NDVI map. We wanted to capture it as a long-term indicator of productivity by using a 3-year growing-season median, which seemed to work well.

*But does NDVI affect the C:N ratio directly or is it more likely to be a proxy for some other property?*

## Limitations

-   WorldClim data from 1970 to 2000 - climate change might already have had an effect by 2018
-   

# AI documentation

## Preparation phase

-   get an overview of possible research directions
-   refine search terms

## Coding

# Appendix



## Spatial autocorrelation C:N

```{r c-n-experimental-variogram}

# use projection (in m)
lucas_proj <- st_transform(lucas, 'EPSG:25832')

# make gstat object
gstat_cn <- gstat(formula = cn_ratio ~ 1, data = lucas_proj)

vg_cn <- variogram(gstat_cn, cloud=F, boundaries=seq(3000, 30000, 3000))
plot(vg_cn, plot.nu = TRUE)

vg_cn <- variogram(gstat_cn, cloud=F, boundaries=seq(2000, 10000, 2000))
plot(vg_cn, plot.nu = TRUE)

```

--\> spatial autocorrelation only to distance of \~7.5km and only small fraction of points have other points within this distance (especially when trying variogram with smaller cloud classes), so geostatistical approaches such as kriging not ideal


## Predictor plots and histograms




# References
