---
title: "Digital soil mapping of C:N ratio across Germany"
author: "Nina Preußler"
format: html
editor: visual
---

# Introduction

## Why is C:N important?

-   What does it tell you about the soil?

## How is C:N changing?

-   What processes and factors affect it?
-   EU/Germany trends

## Why DSM?

-   existing maps for EU/Germany
    -   method
    -   data
    -   quality
    -   challenges

# Methods

```{r load-packages}

library(sf)
library(terra)
library(tmap)
library(ggplot2)
library(stars)
library(tidyverse)
library(devtools)
library(geodata)

```

## Study area

```{r load-germany-mask}

library(rnaturalearth)
library(devtools)
devtools::install_github("ropensci/rnaturalearthhires", quiet = TRUE)

# get Germany boundary polygon
germany <- ne_countries(scale = 10, country = "Germany", returnclass = "sf") %>%
  select(geometry)
```

## Data

## LUCAS soil survey

LUCAS topsoil survey 2018

```{r load-lucas-data}
setwd('~/Documents/00_ESDSRS/semester1/3_soil/soil-N-project/data')

# load point data
lucas_eu <- st_read('LUCAS-SOIL-2018-v2/LUCAS-SOIL-2018.shp')

# load attribute data
lucas_data <- read.csv('LUCAS-SOIL-2018-v2/LUCAS-SOIL-2018.csv')

# merge
lucas_eu <- lucas_eu %>%
  inner_join(lucas_data, by = join_by(POINTID))

# subset to Germany
lucas <- st_filter(lucas_eu, germany)

lucas %>% glimpse()
plot(lucas["N"])
```

### C:N ratio

calculated with $\frac{\text{total N}}{\text{organic C}}$

```{r calculate-cn-ratio}

# first make columns numeric, then add new column with ratio
lucas <- lucas %>%
  mutate(N = if_else(N == "< LOD", 0, as.numeric(N)),
         OC = if_else(OC == "< LOD", 0, as.numeric(OC)),
         cn_ratio = OC / N)

ggplot(lucas, aes(cn_ratio)) +
  geom_histogram(binwidth = 0.5)
```

```{r remove-nans-and-outliers}

# copy before cleaning
lucas_init <- lucas

lucas <- lucas %>%
  drop_na(cn_ratio)

cn_iqr <- IQR(lucas$cn_ratio)
cn_mean <- mean(lucas$cn_ratio)
cn_median <- median(lucas$cn_ratio)
cn_sd <- sd(lucas$cn_ratio)

```

```{r plot-lucas-points-germany}

tmap_mode("view")
tm_basemap('Esri.WorldImagery') +
  tm_shape(germany) +
  tm_lines(col="black", lwd = 3) +
  tm_shape(lucas) +
  tm_dots(fill="cn_ratio",
          size = 0.5,
          fill.scale = tm_scale_continuous(values = "cividis",
                                           limits = c(0, 20)))
```

#### Distance between sampling points

```{r}
library(spatstat)

# prepare point pattern with projection (required for distance)
lucas_ppp <- lucas %>%
  st_transform('EPSG:25832') %>%
  as.ppp()

# find distance to nearest neighbor for each point
nncross(lucas_ppp, lucas_ppp, what='dist')
```

## Covariates

based on Khosravani et al. (2025)

| Abbreviation | Full name | Unit | Source |
|--------------|-----------|------|--------|
|      ann_temp        |    Mean annual temperature       |      |        |
|       ann_precip       |           |      |        |
|     slope         |      Slope     |      |   https://zenodo.org/records/1447210     |
|              |           |      |        |
|              |           |      |        |
|              |           |      |        |
|              |           |      |        |
|              |           |      |        |
|              |           |      |        |
|              |           |      |        |
|              |           |      |        |
|              |           |      |        |
|              |           |      |        |
|              |           |      |        |

: Covariates used in C:N prediction, as proposed by Khosravani et al. (2025)


### Temperature and precipitation

- loaded from WorldClim (XXX)

```{r load-worldclim-data}

library(geodata)

# WorldClim (temperature and precipitation)
temp <- worldclim_country(country='germany', var='tavg', path='data/covariates/worldclim', res=0.5)
precip <- worldclim_country(country='germany', var='prec', path='data/covariates/worldclim', res=0.5)

temp_annual <- mean(temp) # mean annual temperature
precip_annual <- sum(precip) # total annual precipitation

```


### Slope and aspect

Used DEM from Hengl /  Yamazaki et al (https://zenodo.org/records/1447210).

```{r load-dem}

setwd('/Users/ninap/Documents/00_ESDSRS/semester1/3_soil/soil-N-project/data/covariates/elevation')

dem_global <- rast('dtm_elevation_merit.dem_m_250m_s0..0cm_2017_v1.0.tif')

# crop to Germany
dem <- crop(dem_global, germany)

```

```{r calculate-slope-and-aspect}

# Calculate aspect (results in degrees by default)
slope <- terrain(dem, v = "slope", unit = "degrees")

# Calculate aspect (results in degrees by default)
aspect <- terrain(dem, v = "aspect", unit = "degrees")

```

calculated slope and aspect in degrees using terra

### Latitude

extracted latitude from rasters
```{r}
rast_coords <- rast(list(lon = init(dem, "x"), lat = init(dem, "y")))
plot(rast_coords)
```

```{r}
latitude <- rast(list(latitude = init(dem, "y")))
plot(latitude)
```

```{r}

rast_coords <- rast(dem, nlyrs=2)
names(rast_coords) <- c("lon", "lat")
values(rast_coords) <- crds(dem)

# Plot the X and Y coordinates as rasters
plot(rast_coords)

```



### Microbial biomass

```{r load-microbial-biomass-data}
setwd('~/Documents/00_ESDSRS/semester1/3_soil/soil-N-project/data/GEB_maps')

microb <- rast('Cmic_gen_annual_mean_t.tif')
microb_eu <- read_stars('Cmic_gen_annual_mean_t.tif')

# subset to
#microb <- st_filter(microb_eu, germany)

plot(microb_eu)
```

## Random Forest model

## Model evaluation

-   coefficient of determination (R2)
-   concordance correlation coefficient (CCC)
-   root mean squared error (RMSE)
-   Bias

### Uncertainty map

# Results

## Exploratory data analysis

## Evalutation

### Summary statistics of predicted C:N by land cover class

# Discussion

## Limitations

- WorldClim data from 1970 to 2000 - climate change might already have had an effect by 2018

### Microbial

------------------------------------------------------------------------

# AI documentation

## Preparation phase

-   get an overview of possible research directions
-   refine search terms
