---
title: "Digital soil mapping of C:N ratio across Germany"
author: "Nina Preußler"
format: html
editor: visual
---

# Introduction

### Why is C:N important?
- What does it tell you about the soil?

### How is C:N changing?
- What processes and factors affect it?
- EU/Germany trends

### DSM
- existing maps for EU/Germany
  - method
  - data
  - quality
  - challenges

# Methods

```{r load-packages}

library(sf)
library(terra)
library(tmap)
library(ggplot2)
library(stars)
library(tidyverse)

```

## Study area

```{r load-germany-mask}

library(rnaturalearth)

# get Germany boundary polygon
germany <- ne_countries(scale = 10, country = "Germany", returnclass = "sf") %>%
  select(geometry)
```

## Data

### C:N ratio

obtained from LUCAS topsoil survey 2018

```{r load-lucas-data}
setwd('~/Documents/00_ESDSRS/semester1/3_soil/soil-N-project/data')

# load point data
lucas_eu <- st_read('LUCAS-SOIL-2018-v2/LUCAS-SOIL-2018.shp')

# load attribute data
lucas_data <- read.csv('LUCAS-SOIL-2018-v2/LUCAS-SOIL-2018.csv')

# merge
lucas_eu <- lucas_eu %>% 
  inner_join(lucas_data, by = join_by(POINTID))

# subset to Germany
lucas <- st_filter(lucas_eu, germany)

lucas %>% glimpse()
plot(lucas$N)

```


calculated with $\frac{\text{total N}}{\text{organic C}}$

```{r}
lucas <- lucas %>%
  # first make columns numeric, then add new column with ratio
  mutate(N = if_else(N == "< LOD", 0, as.numeric(N)),
         OC = if_else(OC == "< LOD", 0, as.numeric(OC)),
         cn_ratio = OC / N) 

ggplot(lucas, aes(cn_ratio)) +
  geom_histogram(binwidth = 0.5)
```

```{r plot-lucas-points-germany}

tmap_mode("view")
tm_basemap('Esri.WorldImagery') +
  tm_shape(germany) +
  tm_lines(col="black", lwd = 3) +
  tm_shape(lucas) +
  tm_dots(fill="cn_ratio", 
          size = 0.5,   
          fill.scale = tm_scale_continuous(values = "cividis",
                                           limits = c(0, 20)))
```

##### Distance between sampling points

```{r}
library(spatstat)

#nncross(lucas, lucas)
```


### Microbial biomass

```{r}
setwd('~/Documents/00_ESDSRS/semester1/3_soil/soil-N-project/data/GEB_maps')

microb <- rast('Cmic_gen_annual_mean_t.tif')
microb_eu <- read_stars('Cmic_gen_annual_mean_t.tif')

# subset to 
#microb <- st_filter(microb_eu, germany)

plot(microb_eu)
```

# Results

## Exploratory data analysis

# AI documentation

## Preparation phase

-   get an overview of possible research directions
-   refine search terms
